%% SCRAP: experiments/campaigns/l8_attractor_map/SESSION_REPORT_2025-12-10
%% SOURCE: docs/working/experiments/campaigns/l8_attractor_map/SESSION_REPORT_2025-12-10.md
%% STATUS: HISTORICAL
%% FITS: experiments/ch-l8-map
%% EDITORIAL: lifted — prose rewritten to press voice

\section{L8 Attractor Map: Session Report, 10 December 2025}

This session report documents a six-hour experimental session that began
as a memory-safety debugging exercise and ended with the identification of
potentially fundamental computational-physics behavior. The empirical
findings are current; the theoretical interpretations remain under
validation. The 38\,580-run empirical basis (38\,400 DoE + 180 attractor
runs) is the established fact; broader theoretical claims await
cross-platform replication.

\subsection{Memory Safety Fixes}

The L8 attractor experiment exhibited a 35--40\% non-deterministic crash
rate. AddressSanitizer and UndefinedBehaviorSanitizer identified four
bugs:

\begin{description}
  \item[Bug 1 — Heap-buffer-overflow in \texttt{dictionary\_heat\_optimization.c}.]
        A race condition caused \texttt{sf\_fc\_count[bucket\_id]} to
        exceed \texttt{sf\_fc\_cap[bucket\_id]}. Fix: exposed the capacity
        array and added a bounds check with TOCTOU protection before the
        \texttt{memcpy} in \texttt{dict\_reorganize\_buckets\_by\_heat}.
  \item[Bug 2 — Use-after-free in \texttt{dictionary\_heat\_optimization.c}.]
        The heartbeat thread sorted dictionary entry pointers while the
        main thread freed entries. Fix: acquire \texttt{dict\_lock} for
        the entire bucket reorganization pass.
  \item[Bug 3 — UBSan left-shift of negative value in M/MOD.]
        Left-shifting a negative signed value is undefined behavior in C99.
        Fix: cast the dividend to \texttt{unsigned long long} before
        shifting, then cast back.
  \item[Bug 4 — UBSan left-shift in Q48.16 conversion.]
        Fix: replace left-shift with multiplication for signed delta values
        in \texttt{src/doe\_metrics.c}.
\end{description}

After fixes: 50 consecutive ASan runs, zero errors; 100 production runs,
zero crashes; 180-run L8 experiment, 100\% success.

\subsection{Experimental Infrastructure}

\subsubsection{Workload Identification}

Because runs were randomized (to eliminate temporal bias), the heartbeat
CSV carried no workload metadata. The solution cross-referenced the run
matrix against VM restart markers: heartbeat ticks with
\texttt{tick\_interval\_ns} $> 10^9$ indicate VM startup. This yielded a
\texttt{workload\_mapping.csv} cross-reference table without breaking
randomization.

\subsubsection{Analysis Pipeline}

\begin{lstlisting}[language=bash]
cd experiments/l8_attractor_map
./scripts/setup.sh   # Create structure, CSV headers
./scripts/run.sh     # Execute 180 runs, generate mapping
Rscript l8_analysis_safe.R  # Statistics + 13 SVG charts
\end{lstlisting}

\subsection{Statistical Results}

180 runs (6 workloads $\times$ 30 replicates), 4\,171 valid heartbeat
ticks after filtering.

\begin{center}
\begin{tabular}{llll}
\toprule
Workload    & Mean ($\mu$s) & CV (\%) & Character \\
\midrule
STABLE      & 75.45 & 18.00 & Lowest variability \\
VOLATILE    & 76.04 & 22.71 & Moderate \\
OMNI        & 76.95 & 26.20 & 7$\times$ compute, comparable stability \\
TEMPORAL    & 75.77 & 32.09 & Moderate \\
DIVERSE     & 76.87 & 42.47 & High variability \\
TRANSITION  & 81.72 & 69.13 & Highest variability (phase-change region) \\
\bottomrule
\end{tabular}
\end{center}

ANOVA on convergence time: $F(5,174) = 0.983$, $p = 0.43$ — convergence
time is workload-independent, confirming deterministic behavior.
ANOVA on tick intervals: $F(5,4165) = 3.890$, $p = 0.0016$ — workloads
produce statistically distinct timing signatures.

\subsection{Observed Phenomena}

All six workloads converged to a ground-state oscillation frequency of
approximately $\omega_0 \approx 13.5$\,Hz (CV 1.3\% across workloads).
The damped harmonic oscillator equation $\omega(t) = \omega_0 +
A e^{-\gamma t} \cos(\Omega t + \phi)$ fit the convergence trajectories.
Each workload exhibited a distinct effective temperature $T_\text{eff}$
under a Boltzmann-distribution model of frequency fluctuations
(STABLE: $T_\text{eff} = 2.175$\,Hz; TRANSITION: $T_\text{eff} =
2.691$\,Hz). Phase-space portraits revealed 45-degree diagonal
relationships across variable pairs, suggesting conserved quantities.

Probability assessments from the session:
\begin{itemize}
  \item $> 90\%$ confidence: deterministic convergence, workload-independent
        convergence time, reproducible CV as a metric.
  \item 70--85\% confidence: $\omega_0 \approx 13.5$\,Hz is a real pattern;
        adaptive window finds equilibrium $W^*$; James Law $K \approx 1.0$;
        workload spectral signatures are distinct and stable.
  \item $< 30\%$ confidence: $\omega_0$ is a true universal constant;
        hardware independence; general computational law.
\end{itemize}

The frequency likely synchronizes with the CPU clock and is therefore a
hardware-dependent emergent property rather than a universal constant.

\subsection{The Adaptive Window Revelation}

A critical finding from this session: \texttt{effective\_window\_size} is
\emph{not} logged in the heartbeat CSV. The adaptive window
(\texttt{src/rolling\_window\_of\_truth.c}) shrinks to 75\% when pattern
diversity growth falls below 1\%, and grows back toward
\texttt{ROLLING\_WINDOW\_SIZE} otherwise. The system finds an equilibrium
$W^*$ at which it ``flatlines.'' This equilibrium was empirically observed
but not yet instrumented; adding \texttt{effective\_window\_size} to the
heartbeat CSV was identified as Priority 1 for the next session.

\subsection{James Law}

Empirical observation from this session: $K = \Lambda \times (\text{DoF}+1)
/ W \approx 1.0$, where $\Lambda$ is the effective smoothing factor,
$W$ the window size, and DoF the number of active feedback loops. This
relationship was named James Law and is under empirical validation in the
window-scaling campaign.

%% PATENT: James Law ($\Lambda = W / (\text{DoF}+1)$) and the
%% spectroscopic workload-fingerprinting concept are patent-adjacent.
%% Do not draft claims here.

\subsection{Next Validation Steps}

\begin{enumerate}
  \item Add \texttt{effective\_window\_size} logging to heartbeat CSV
        and re-run 180-run L8 experiment to observe $W^*$ trajectory.
  \item Run the window-scaling experiment (2\,880 runs; see companion
        README) to validate James Law across 8 DoF $\times$ 12 window
        sizes.
  \item Cross-platform validation on ARM to test whether $\omega_0$
        varies with CPU architecture.
\end{enumerate}
