%% User Guide — main.tex
%% SOURCE: docs/working/architecture/getting-started/, docs/working/architecture/architecture-internals/
%% TODO(bob): set version, ISBN/DOI placeholder, cover design

\input{../common/preamble}

%% longtable for the word reference appendix
\usepackage{longtable}
\usepackage{array}

\title{StarForth User Guide\\[0.4em]
       \large Running, Scripting, and Exploring StarForth}
\author{Robert A.\ James}
\date{%% TODO(bob): set edition date
}

\begin{document}
\frontmatter
\maketitle

\tableofcontents

%%────────────────────────────────────────────────────────────────────────────
\chapter*{About This Guide}
%%────────────────────────────────────────────────────────────────────────────

This guide covers StarForth from the perspective of someone sitting at a
terminal running it. You might know FORTH already, or this might be your first
encounter with stack-based programming. Either way, the goal is the same: get
you productive at the prompt.

StarForth is a FORTH-79 compliant virtual machine that also happens to contain
a physics-driven adaptive runtime. Most of the time the runtime is invisible ---
it tunes itself in the background while you type. When you want to inspect or
control it, this guide covers that too.

The guide is organized so you can read it front-to-back or jump to the chapter
you need. Part~I gets you running. Part~II through Part~VI cover the language.
Part~VII covers the physics features unique to StarForth. Part~VIII gives a
brief orientation to the access-control system. The appendix is a compact
word-reference table you can keep open alongside your terminal.

\mainmatter

%%━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━
\part{First Steps}
%%━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━

\chapter{Installation}
\input{../scraps/architecture/getting-started/INSTALL}

\chapter{Quick Start}
\input{../scraps/architecture/getting-started/QUICKSTART}

%%────────────────────────────────────────────────────────────────────────────
\chapter{Your First Session}
\label{chap:first-session}
%%────────────────────────────────────────────────────────────────────────────

This chapter walks through a complete interactive session from starting the
REPL to shutting it down. Every prompt and response is annotated.

%% ─────────────────────────────────────────────────────────────────────────
\section{Starting StarForth}
\label{sec:starting}

After building, start the interactive REPL:

\begin{lstlisting}[language=bash]
./build/amd64/standard/starforth
\end{lstlisting}

StarForth loads \texttt{init.4th}, sets up the dictionary, and prints the
prompt:

\begin{lstlisting}[language=Forth]
ok>
\end{lstlisting}

The \texttt{ok>} prompt means the system is ready, the data stack is empty, and
you are in interpretation mode. Everything you type is interpreted immediately.
There is no compile-then-run cycle for interactive work.

%% ─────────────────────────────────────────────────────────────────────────
\section{Pushing Numbers}
\label{sec:push-numbers}

Type a number and press Enter. StarForth pushes it onto the data stack:

\begin{lstlisting}[language=Forth]
ok> 42
ok>
\end{lstlisting}

No output yet --- the number is on the stack waiting to be used. Push another:

\begin{lstlisting}[language=Forth]
ok> 7
ok>
\end{lstlisting}

Now two numbers are on the stack: 42 on the bottom, 7 on top.

%% ─────────────────────────────────────────────────────────────────────────
\section{Arithmetic}
\label{sec:arithmetic-first}

The word \texttt{+} pops the top two items, adds them, and pushes the result:

\begin{lstlisting}[language=Forth]
ok> 42 7 +
ok>
\end{lstlisting}

Still no output --- the result (49) is sitting on the stack. Use the dot word
\texttt{.} to pop and print it:

\begin{lstlisting}[language=Forth]
ok> 42 7 + .
49 ok>
\end{lstlisting}

You can write this all on one line. StarForth reads left to right, executing
each word as it encounters it:

\begin{lstlisting}[language=Forth]
ok> 3 4 * .
12 ok>
\end{lstlisting}

Multiplication works the same way. All arithmetic is postfix: the operands
come before the operator.

%% ─────────────────────────────────────────────────────────────────────────
\section{Inspecting the Stack with .S}
\label{sec:dot-s}

\texttt{.S} (dot-S) prints the entire stack without consuming anything. This
is your main tool for understanding what is happening:

\begin{lstlisting}[language=Forth]
ok> 10 20 30
ok> .S
<3> 10 20 30 ok>
\end{lstlisting}

The \texttt{<3>} tells you there are three items. The items are shown
bottom-to-top, so the rightmost number (30) is on top. The stack is
unchanged after \texttt{.S}.

Clear the stack with \texttt{DROP} repeated, or define a word later, or just
exit and restart. For now:

\begin{lstlisting}[language=Forth]
ok> DROP DROP DROP
ok> .S
<0> ok>
\end{lstlisting}

%% ─────────────────────────────────────────────────────────────────────────
\section{Listing Available Words with WORDS}
\label{sec:words}

\texttt{WORDS} prints every word currently in the dictionary:

\begin{lstlisting}[language=Forth]
ok> WORDS
WORD-ENTROPY PHYSICS-FREEZE PHYSICS-THAW WORDS .S .
DUP DROP SWAP ROT OVER NIP TUCK 2DUP 2DROP 2SWAP
... (many more) ...
ok>
\end{lstlisting}

The list is long. StarForth ships the full FORTH-79 word set plus
extensions. Use \texttt{WORDS} when you cannot remember a word's exact name.

%% ─────────────────────────────────────────────────────────────────────────
\section{Defining Your First Word}
\label{sec:first-definition}

Use \texttt{:} to define a new word and \texttt{;} to end the definition:

\begin{lstlisting}[language=Forth]
ok> : DOUBLE  DUP + ;
ok> 7 DOUBLE .
14 ok>
\end{lstlisting}

\texttt{DOUBLE} is now in the dictionary. It takes one number, duplicates it,
and adds the two copies together. You can use it anywhere.

%% ─────────────────────────────────────────────────────────────────────────
\section{Errors and Recovery}
\label{sec:errors}

Type something StarForth does not recognize:

\begin{lstlisting}[language=Forth]
ok> FOO
Error: word not found: FOO
ok>
\end{lstlisting}

StarForth reports the error and returns to the prompt. The stack is reset to
the state it was in before the offending line. You can keep working:

\begin{lstlisting}[language=Forth]
ok> 3 4 + .
7 ok>
\end{lstlisting}

Stack underflow (trying to pop from an empty stack) is also caught:

\begin{lstlisting}[language=Forth]
ok> .S
<0> ok>
ok> DROP
Error: stack underflow
ok>
\end{lstlisting}

The system does not crash. Error recovery brings you back to a clean prompt.

%% ─────────────────────────────────────────────────────────────────────────
\section{Exiting}
\label{sec:bye}

Type \texttt{BYE} to exit:

\begin{lstlisting}[language=Forth]
ok> BYE
\end{lstlisting}

The VM shuts down cleanly. Any words you defined in the session are gone;
nothing is saved automatically. To save work, write it to a source file and
load it next time with \texttt{INCLUDE} (see Chapter~\ref{chap:files}).

%% ─────────────────────────────────────────────────────────────────────────
\section{Command-Line Shortcuts}
\label{sec:cli}

You do not have to run interactively. You can pass code directly:

\begin{lstlisting}[language=bash]
./build/amd64/standard/starforth -c "3 4 + . BYE"
7
\end{lstlisting}

Run the test suite and then drop into the REPL:

\begin{lstlisting}[language=bash]
./build/amd64/standard/starforth --run-tests
\end{lstlisting}

%%━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━
\part{The Stack}
%%━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━

\chapter{Stack Fundamentals}
\label{chap:stack-fundamentals}

FORTH is a stack-based language. Almost everything you do involves the data
stack. Understanding it well makes everything else easier.

%% ─────────────────────────────────────────────────────────────────────────
\section{What the Stack Is}
\label{sec:what-the-stack-is}

The data stack is a last-in, first-out (LIFO) sequence of cells. A cell in
StarForth is a 64-bit signed integer on a 64-bit host. You push values onto
the top of the stack and pop them from the top.

Numbers push automatically. Words consume values (pop) and may produce values
(push). The stack effect notation describes what a word does:

\begin{center}
\texttt{( before -- after )}
\end{center}

Items before the \texttt{--} are consumed (top of stack on the right). Items
after are produced. For example, \texttt{+} has the stack effect
\texttt{( a b -- a+b )}: it pops \texttt{b}, pops \texttt{a}, and pushes
their sum.

By convention in this guide, the topmost item is labeled \texttt{n} or
\texttt{a}; the next item down is \texttt{n2} or \texttt{b}; and so on.

%% ─────────────────────────────────────────────────────────────────────────
\section{Stack Depth}
\label{sec:stack-depth}

The default data stack depth is 256 cells (configurable at build time via
\texttt{DATA\_STACK\_SIZE}). Pushing more than 256 items causes a stack
overflow error. Popping from an empty stack causes a stack underflow error.
Both are caught and reported; neither crashes the system.

%% ─────────────────────────────────────────────────────────────────────────
\section{The Data Stack vs.\ the Return Stack}
\label{sec:two-stacks}

StarForth maintains two stacks:

\begin{description}
  \item[Data stack] The main working stack. All arithmetic and most word
    operations use it. This is the one you interact with directly.
  \item[Return stack] Manages word call/return linkage and is used as
    temporary storage with \texttt{>R}, \texttt{R>}, and \texttt{R@}.
    The return stack is 256 cells deep by default. Do not leave items on the
    return stack across word boundaries --- that causes unpredictable
    behavior.
\end{description}

%%────────────────────────────────────────────────────────────────────────────
\chapter{Core Stack Words}
\label{chap:core-stack}
%%────────────────────────────────────────────────────────────────────────────

This chapter covers the words that manipulate the data stack directly.

%% ─────────────────────────────────────────────────────────────────────────
\section{Single-Item Operations}
\label{sec:single-stack}

\subsection*{DUP \texttt{( n -- n n )}}

Duplicates the top of the stack.

\begin{lstlisting}[language=Forth]
ok> 5 DUP .S
<2> 5 5 ok>
\end{lstlisting}

\subsection*{DROP \texttt{( n -- )}}

Removes the top of the stack.

\begin{lstlisting}[language=Forth]
ok> 5 7 DROP .S
<1> 5 ok>
\end{lstlisting}

\subsection*{SWAP \texttt{( a b -- b a )}}

Exchanges the top two items.

\begin{lstlisting}[language=Forth]
ok> 1 2 SWAP .S
<2> 2 1 ok>
\end{lstlisting}

\subsection*{ROT \texttt{( a b c -- b c a )}}

Rotates the top three items: the third item from the top comes to the top.

\begin{lstlisting}[language=Forth]
ok> 1 2 3 ROT .S
<3> 2 3 1 ok>
\end{lstlisting}

\subsection*{OVER \texttt{( a b -- a b a )}}

Copies the second item to the top, leaving the original in place.

\begin{lstlisting}[language=Forth]
ok> 10 20 OVER .S
<3> 10 20 10 ok>
\end{lstlisting}

\subsection*{NIP \texttt{( a b -- b )}}

Removes the second item (drops below top).

\begin{lstlisting}[language=Forth]
ok> 10 20 NIP .S
<1> 20 ok>
\end{lstlisting}

\subsection*{TUCK \texttt{( a b -- b a b )}}

Copies the top item below the second item.

\begin{lstlisting}[language=Forth]
ok> 10 20 TUCK .S
<3> 20 10 20 ok>
\end{lstlisting}

%% ─────────────────────────────────────────────────────────────────────────
\section{Double-Item Operations}
\label{sec:double-stack}

These words operate on pairs of items (a ``double''), treating two adjacent
cells as a single logical unit.

\subsection*{2DUP \texttt{( a b -- a b a b )}}

Duplicates the top pair.

\begin{lstlisting}[language=Forth]
ok> 1 2 2DUP .S
<4> 1 2 1 2 ok>
\end{lstlisting}

\subsection*{2DROP \texttt{( a b -- )}}

Drops the top pair.

\begin{lstlisting}[language=Forth]
ok> 1 2 3 4 2DROP .S
<2> 1 2 ok>
\end{lstlisting}

\subsection*{2SWAP \texttt{( a b c d -- c d a b )}}

Exchanges the top two pairs.

\begin{lstlisting}[language=Forth]
ok> 1 2 3 4 2SWAP .S
<4> 3 4 1 2 ok>
\end{lstlisting}

%% ─────────────────────────────────────────────────────────────────────────
\section{The Return Stack}
\label{sec:return-stack}

The return stack is normally managed automatically by StarForth's word-call
machinery. You can borrow it as temporary storage within a single word
definition using these three words.

\subsection*{>R \texttt{( n -- ) (R: -- n )}}

Moves the top of the data stack to the top of the return stack.

\subsection*{R> \texttt{( -- n ) (R: n -- )}}

Moves the top of the return stack to the top of the data stack.

\subsection*{R@ \texttt{( -- n ) (R: n -- n )}}

Copies (does not move) the top of the return stack to the data stack.

A typical use: stash a loop counter temporarily while performing a computation.

\begin{lstlisting}[language=Forth]
: EXAMPLE  ( a b -- result )
    >R       \ stash b on return stack
    DUP *    \ square a
    R>       \ retrieve b
    + ;      \ add b
\end{lstlisting}

\textbf{Rule}: every \texttt{>R} in a word must be matched by an \texttt{R>}
before the word exits, and you must not use \texttt{>R}/\texttt{R>} to move
items across loop boundaries (DO LOOP manages the return stack internally).

%%━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━
\part{Arithmetic and Logic}
%%━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━

\chapter{Integer Arithmetic}
\label{chap:arithmetic}

%% ─────────────────────────────────────────────────────────────────────────
\section{Basic Arithmetic Words}
\label{sec:basic-arith}

\subsection*{+ \texttt{( a b -- a+b )}}

Adds two integers.

\begin{lstlisting}[language=Forth]
ok> 100 23 + .
123 ok>
\end{lstlisting}

\subsection*{- \texttt{( a b -- a-b )}}

Subtracts: removes \texttt{b} from \texttt{a}.

\begin{lstlisting}[language=Forth]
ok> 100 23 - .
77 ok>
\end{lstlisting}

\subsection*{* \texttt{( a b -- a*b )}}

Multiplies.

\begin{lstlisting}[language=Forth]
ok> 7 8 * .
56 ok>
\end{lstlisting}

\subsection*{/ \texttt{( a b -- a/b )}}

Integer division; truncates toward zero.

\begin{lstlisting}[language=Forth]
ok> 17 5 / .
3 ok>
\end{lstlisting}

\subsection*{MOD \texttt{( a b -- a mod b )}}

Remainder after division.

\begin{lstlisting}[language=Forth]
ok> 17 5 MOD .
2 ok>
\end{lstlisting}

\subsection*{ABS \texttt{( n -- |n| )}}

Absolute value.

\begin{lstlisting}[language=Forth]
ok> -42 ABS .
42 ok>
\end{lstlisting}

\subsection*{MIN \texttt{( a b -- min )}}

Pushes the smaller of two values.

\begin{lstlisting}[language=Forth]
ok> 10 20 MIN .
10 ok>
\end{lstlisting}

\subsection*{MAX \texttt{( a b -- max )}}

Pushes the larger of two values.

\begin{lstlisting}[language=Forth]
ok> 10 20 MAX .
20 ok>
\end{lstlisting}

%% ─────────────────────────────────────────────────────────────────────────
\section{Comparison Words}
\label{sec:comparison}

Comparisons push a flag: \texttt{-1} (all bits set) for true, \texttt{0} for
false. This is the standard FORTH-79 flag convention.

\begin{center}
\begin{tabular}{lll}
\toprule
Word & Stack effect & Meaning \\
\midrule
\texttt{=}  & \texttt{( a b -- flag )} & equal \\
\texttt{<>} & \texttt{( a b -- flag )} & not equal \\
\texttt{<}  & \texttt{( a b -- flag )} & a less than b \\
\texttt{>}  & \texttt{( a b -- flag )} & a greater than b \\
\texttt{<=} & \texttt{( a b -- flag )} & a less than or equal to b \\
\texttt{>=} & \texttt{( a b -- flag )} & a greater than or equal to b \\
\texttt{0=} & \texttt{( n -- flag )} & true if n is zero \\
\texttt{0<} & \texttt{( n -- flag )} & true if n is negative \\
\texttt{0>} & \texttt{( n -- flag )} & true if n is positive \\
\bottomrule
\end{tabular}
\end{center}

\begin{lstlisting}[language=Forth]
ok> 5 3 > .
-1 ok>
ok> 5 3 < .
0 ok>
\end{lstlisting}

%%────────────────────────────────────────────────────────────────────────────
\chapter{Logical and Bitwise Operations}
\label{chap:logic}
%%────────────────────────────────────────────────────────────────────────────

%% ─────────────────────────────────────────────────────────────────────────
\section{Bitwise Words}
\label{sec:bitwise}

In StarForth, the logical words operate bitwise on cells. Because FORTH flags
are all-bits-set (\texttt{-1}) for true and zero for false, bitwise AND and OR
also work correctly as logical AND and OR when applied to flags.

\subsection*{AND \texttt{( a b -- a\&b )}}

Bitwise AND.

\begin{lstlisting}[language=Forth]
ok> 0b1100 0b1010 AND .
8 ok>     ( 0b1000 = 8 )
\end{lstlisting}

\subsection*{OR \texttt{( a b -- a|b )}}

Bitwise OR.

\begin{lstlisting}[language=Forth]
ok> 0b1100 0b1010 OR .
14 ok>    ( 0b1110 = 14 )
\end{lstlisting}

\subsection*{XOR \texttt{( a b -- a xor b )}}

Bitwise exclusive OR.

\begin{lstlisting}[language=Forth]
ok> 0b1100 0b1010 XOR .
6 ok>     ( 0b0110 = 6 )
\end{lstlisting}

\subsection*{INVERT \texttt{( n -- $\sim$n )}}

Bitwise complement of all bits.

\begin{lstlisting}[language=Forth]
ok> 0 INVERT .
-1 ok>
\end{lstlisting}

\subsection*{NOT \texttt{( n -- flag )}}

Logical NOT: pushes \texttt{-1} if \texttt{n} is zero, \texttt{0} otherwise.
Equivalent to \texttt{0=}.

\subsection*{LSHIFT \texttt{( n count -- n shifted )}}

Shifts \texttt{n} left by \texttt{count} bits.

\begin{lstlisting}[language=Forth]
ok> 1 4 LSHIFT .
16 ok>
\end{lstlisting}

\subsection*{RSHIFT \texttt{( n count -- n shifted )}}

Shifts \texttt{n} right by \texttt{count} bits (logical shift: fills with
zeros).

\begin{lstlisting}[language=Forth]
ok> 256 4 RSHIFT .
16 ok>
\end{lstlisting}

%%────────────────────────────────────────────────────────────────────────────
\chapter{Q48.16 Fixed-Point Arithmetic}
\label{chap:q48-16}
%%────────────────────────────────────────────────────────────────────────────

%% ─────────────────────────────────────────────────────────────────────────
\section{What Q48.16 Is}
\label{sec:q48-what}

\Qtype{} is StarForth's built-in fixed-point number format. A \Qtype{} value
is a 64-bit integer where the lower 16 bits represent the fractional part and
the upper 48 bits represent the integer part. The precision is
$1/65536 \approx 0.0000153$.

You encounter \Qtype{} values primarily through the physics runtime (heat
values, decay slopes) and in words prefixed with \texttt{Q}. In the REPL you
will see them printed as rational numbers or as raw integers depending on the
display word used.

%% ─────────────────────────────────────────────────────────────────────────
\section{When You See Q48.16}
\label{sec:q48-when}

The REPL does not automatically distinguish a \Qtype{} value from a plain
integer --- both are cells. Context tells you which you have. The physics
diagnostics (Chapter~\ref{chap:physics}) print heat and slope values in
\Qtype{} form. If you use \texttt{WORD-ENTROPY}, expect to see heat counters
that may look like large integers; these are \Qtype{} values with the fractional
part encoded in the low 16 bits.

%% ─────────────────────────────────────────────────────────────────────────
\section{Q48.16 Words}
\label{sec:q48-words}

The \texttt{q48\_16\_words.c} module provides the Q48.16 word set. Key words:

\begin{center}
\begin{tabular}{lll}
\toprule
Word & Stack effect & Description \\
\midrule
\texttt{Q+}      & \texttt{( a b -- a+b )} & Add two \Qtype{} values \\
\texttt{Q-}      & \texttt{( a b -- a-b )} & Subtract \\
\texttt{Q*}      & \texttt{( a b -- a*b )} & Multiply (with precision) \\
\texttt{Q/}      & \texttt{( a b -- a/b )} & Divide \\
\texttt{INT>Q}   & \texttt{( n -- q )}     & Convert integer to \Qtype{} \\
\texttt{Q>INT}   & \texttt{( q -- n )}     & Convert \Qtype{} to integer (truncate) \\
\texttt{Q.}      & \texttt{( q -- )}       & Print \Qtype{} as decimal fraction \\
\bottomrule
\end{tabular}
\end{center}

%%━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━
\part{Defining Words}
%%━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━

\chapter{Colon Definitions}
\label{chap:colon}

%% ─────────────────────────────────────────────────────────────────────────
\section{: and ;}
\label{sec:colon-semi}

The colon word \texttt{:} starts a new word definition. Everything that
follows --- until the matching \texttt{;} --- becomes the body of the new
word:

\begin{lstlisting}[language=Forth]
: SQUARE  ( n -- n^2 )
    DUP * ;
\end{lstlisting}

Break this down:
\begin{itemize}
  \item \texttt{:} switches StarForth from interpretation mode to compilation
    mode.
  \item \texttt{SQUARE} is the name of the new word.
  \item \texttt{DUP *} is the body: duplicate the top, then multiply. The
    comment in parentheses is documentation only --- StarForth ignores it.
  \item \texttt{;} ends the definition and switches back to interpretation
    mode.
\end{itemize}

Now use it:

\begin{lstlisting}[language=Forth]
ok> 9 SQUARE .
81 ok>
\end{lstlisting}

%% ─────────────────────────────────────────────────────────────────────────
\section{A Worked Example}
\label{sec:worked-example}

Define a word that computes the sum of squares of two numbers:

\begin{lstlisting}[language=Forth]
: SQUARE   ( n -- n^2 )  DUP * ;

: SUM-OF-SQUARES  ( a b -- a^2+b^2 )
    SQUARE   \ square b (top)
    SWAP     \ bring a to top
    SQUARE   \ square a
    + ;      \ add
\end{lstlisting}

Test:

\begin{lstlisting}[language=Forth]
ok> 3 4 SUM-OF-SQUARES .
25 ok>
\end{lstlisting}

Verify it is in the dictionary:

\begin{lstlisting}[language=Forth]
ok> WORDS
SUM-OF-SQUARES SQUARE ...
ok>
\end{lstlisting}

%% ─────────────────────────────────────────────────────────────────────────
\section{The SMUDGE Flag}
\label{sec:smudge}

While StarForth is compiling a new word (between \texttt{:} and \texttt{;}),
the word is marked as \emph{smudged}: it exists in the dictionary but is not
visible to searches. This prevents a partially-compiled word from being called
accidentally during its own definition (unless you explicitly use recursion via
\texttt{RECURSIVE}). When \texttt{;} executes, it clears the smudge flag and
the word becomes visible.

If compilation is interrupted by an error, the smudged entry may remain. A
subsequent \texttt{FORGET} or a new definition with the same name will clean it
up.

%% ─────────────────────────────────────────────────────────────────────────
\section{IMMEDIATE}
\label{sec:immediate}

Normally, a word defined with \texttt{:} is compiled into word bodies when
encountered in compilation mode. Marking a word \texttt{IMMEDIATE} tells the
compiler to execute it at compile time instead:

\begin{lstlisting}[language=Forth]
: MY-WORD  ... ; IMMEDIATE
\end{lstlisting}

\texttt{IF}, \texttt{ELSE}, \texttt{THEN}, \texttt{DO}, \texttt{LOOP}, and
all the control-flow words are immediate. They do their work (patching
branches, emitting addresses) during compilation so the compiled word runs
efficiently.

%%────────────────────────────────────────────────────────────────────────────
\chapter{Variables, Constants, and CREATE}
\label{chap:variables}
%%────────────────────────────────────────────────────────────────────────────

%% ─────────────────────────────────────────────────────────────────────────
\section{VARIABLE}
\label{sec:variable}

\texttt{VARIABLE} creates a named cell in memory:

\begin{lstlisting}[language=Forth]
VARIABLE COUNTER
\end{lstlisting}

\texttt{COUNTER} is now a word. Executing it pushes the \emph{address} of the
cell (not the value). Use \texttt{@} to read and \texttt{!} to write:

\begin{lstlisting}[language=Forth]
ok> 0 COUNTER !      \ store 0 into COUNTER
ok> COUNTER @  .     \ fetch and print
0 ok>
ok> COUNTER @ 1 + COUNTER !   \ increment
ok> COUNTER @ .
1 ok>
\end{lstlisting}

%% ─────────────────────────────────────────────────────────────────────────
\section{CONSTANT}
\label{sec:constant}

\texttt{CONSTANT} creates a word that pushes a fixed value:

\begin{lstlisting}[language=Forth]
42 CONSTANT ANSWER
\end{lstlisting}

\begin{lstlisting}[language=Forth]
ok> ANSWER .
42 ok>
\end{lstlisting}

Unlike a variable, you cannot change the value of a constant after it is
defined (without redefining it).

%% ─────────────────────────────────────────────────────────────────────────
\section{CREATE and DOES>}
\label{sec:create-does}

\texttt{CREATE} allocates a dictionary entry without specifying behavior.
\texttt{DOES>} adds a behavior that runs when the created word is executed.
Together they build custom defining words --- words that create other words.

A classic example: arrays.

\begin{lstlisting}[language=Forth]
: ARRAY  ( n "name" -- )
    CREATE  CELLS ALLOT
    DOES>   ( index addr -- element-addr )
    SWAP CELLS + ;
\end{lstlisting}

\begin{lstlisting}[language=Forth]
ok> 10 ARRAY MYDATA
ok> 99 3 MYDATA !     \ store 99 at index 3
ok> 3 MYDATA @ .      \ fetch index 3
99 ok>
\end{lstlisting}

The \texttt{DOES>} part becomes the runtime behavior of any word created by
\texttt{ARRAY}. When \texttt{MYDATA} executes, the address of its data area
is pushed first, then the \texttt{DOES>} code runs with that address on the
stack.

%%━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━
\part{Control Flow}
%%━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━

\chapter{Conditionals and Loops}
\label{chap:control}

Control-flow words are immediate: they compile branches and jumps into the
word body. They only work inside word definitions (between \texttt{:} and
\texttt{;}), not in the interactive REPL directly.

%% ─────────────────────────────────────────────────────────────────────────
\section{IF ... ELSE ... THEN}
\label{sec:if-else-then}

\texttt{IF} pops the top of the stack. If it is non-zero (true), the code
between \texttt{IF} and \texttt{ELSE} (or \texttt{THEN}) executes. If it is
zero, the \texttt{ELSE} branch (if present) executes.

Stack effect of \texttt{IF}: \texttt{( flag -- )}.

\begin{lstlisting}[language=Forth]
: SIGN  ( n -- )
    DUP 0>  IF ." positive" ELSE
    DUP 0<  IF ." negative" ELSE
              ." zero"
            THEN
            THEN
    DROP ;
\end{lstlisting}

\begin{lstlisting}[language=Forth]
ok> 5 SIGN
positive ok>
ok> -3 SIGN
negative ok>
ok> 0 SIGN
zero ok>
\end{lstlisting}

Note: \texttt{IF ... ELSE ... THEN} can be nested. Each \texttt{IF} must be
matched by exactly one \texttt{THEN}.

%% ─────────────────────────────────────────────────────────────────────────
\section{DO ... LOOP}
\label{sec:do-loop}

\texttt{DO} starts a counted loop. It pops two items: the limit and the
initial index (limit on top). The loop variable \texttt{I} is available inside
the loop body.

\texttt{LOOP} increments the index by 1 and exits when the index reaches the
limit.

Stack effect of \texttt{DO}: \texttt{( limit start -- )}.
Stack effect of \texttt{I}: \texttt{( -- index )}.

\begin{lstlisting}[language=Forth]
: COUNT-UP  ( limit -- )
    0 DO  I .  LOOP ;
\end{lstlisting}

\begin{lstlisting}[language=Forth]
ok> 5 COUNT-UP
0 1 2 3 4 ok>
\end{lstlisting}

\texttt{+LOOP} increments the index by a specified amount:

\begin{lstlisting}[language=Forth]
: COUNT-BY-TWO  ( limit -- )
    0 DO  I .  2 +LOOP ;
\end{lstlisting}

\begin{lstlisting}[language=Forth]
ok> 10 COUNT-BY-TWO
0 2 4 6 8 ok>
\end{lstlisting}

For nested loops, \texttt{J} gives the outer loop index.

\texttt{LEAVE} exits a DO loop immediately:

\begin{lstlisting}[language=Forth]
: FIND-FIVE  ( limit -- )
    0 DO
        I 5 = IF  ." found 5 at index " I . CR  LEAVE  THEN
    LOOP ;
\end{lstlisting}

%% ─────────────────────────────────────────────────────────────────────────
\section{BEGIN ... UNTIL}
\label{sec:begin-until}

\texttt{BEGIN} marks the start of a loop. \texttt{UNTIL} pops a flag; if it
is non-zero (true), the loop exits; if zero, control returns to \texttt{BEGIN}.

The loop always executes at least once.

\begin{lstlisting}[language=Forth]
: COUNTDOWN  ( n -- )
    BEGIN
        DUP .
        1 -
        DUP 0=
    UNTIL
    DROP ;
\end{lstlisting}

\begin{lstlisting}[language=Forth]
ok> 5 COUNTDOWN
5 4 3 2 1 ok>
\end{lstlisting}

%% ─────────────────────────────────────────────────────────────────────────
\section{BEGIN ... WHILE ... REPEAT}
\label{sec:begin-while-repeat}

\texttt{WHILE} is a mid-loop test. \texttt{WHILE} pops a flag; if the flag is
zero (false), the loop exits (jumping to after \texttt{REPEAT}); if non-zero,
the loop body continues.

\begin{lstlisting}[language=Forth]
: PRINT-WHILE-POSITIVE  ( n -- )
    BEGIN
        DUP 0>       \ test: is n positive?
    WHILE
        DUP .        \ print n
        1 -          \ decrement
    REPEAT
    DROP ;
\end{lstlisting}

\begin{lstlisting}[language=Forth]
ok> 4 PRINT-WHILE-POSITIVE
4 3 2 1 ok>
\end{lstlisting}

%% ─────────────────────────────────────────────────────────────────────────
\section{A Note on Interactive Compilation}
\label{sec:note-interactive-compile}

Control-flow words like \texttt{IF} and \texttt{DO} compile branch addresses
that only make sense inside a complete word definition. Typing them directly
at the \texttt{ok>} prompt will produce an error. Always wrap control flow
inside \texttt{: ... ;}.

%%━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━
\part{Memory and Blocks}
%%━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━

\chapter{Memory Access}
\label{chap:memory}

%% ─────────────────────────────────────────────────────────────────────────
\section{Cell-Sized Access: @ and !}
\label{sec:at-store}

\texttt{@} (fetch) reads one cell (64 bits on a 64-bit host) from an address.
\texttt{!} (store) writes one cell to an address.

\subsection*{@ \texttt{( addr -- n )}}

Fetch the cell at \texttt{addr}:

\begin{lstlisting}[language=Forth]
ok> VARIABLE X
ok> 42 X !
ok> X @ .
42 ok>
\end{lstlisting}

\subsection*{! \texttt{( n addr -- )}}

Store \texttt{n} at \texttt{addr}:

\begin{lstlisting}[language=Forth]
ok> 99 X !
ok> X @ .
99 ok>
\end{lstlisting}

%% ─────────────────────────────────────────────────────────────────────────
\section{Byte-Sized Access: C@ and C!}
\label{sec:cat-cstore}

\texttt{C@} and \texttt{C!} read and write a single byte (character).

\subsection*{C@ \texttt{( addr -- char )}}

Fetch one byte.

\subsection*{C! \texttt{( char addr -- )}}

Store one byte.

These are used for string manipulation and working with character buffers.

%% ─────────────────────────────────────────────────────────────────────────
\section{Bulk Memory: MOVE and FILL}
\label{sec:move-fill}

\subsection*{MOVE \texttt{( src dst count -- )}}

Copies \texttt{count} bytes from \texttt{src} to \texttt{dst}. Handles
overlapping regions correctly.

\subsection*{FILL \texttt{( addr count char -- )}}

Fills \texttt{count} bytes starting at \texttt{addr} with the byte value
\texttt{char}.

%%────────────────────────────────────────────────────────────────────────────
\chapter{The Block Subsystem}
\label{chap:blocks}
%%────────────────────────────────────────────────────────────────────────────

%% ─────────────────────────────────────────────────────────────────────────
\section{What Blocks Are}
\label{sec:blocks-what}

FORTH-79 uses a block-oriented storage model. A block is 1024 bytes of data,
addressed by a logical block number (LBN). Blocks provide a uniform interface
to storage whether the backing medium is a file, RAM, or physical disk.

In StarForth the block subsystem maps logical block numbers to physical storage
through a pluggable backend (file, RAM, or factory). The default configuration
uses RAM blocks for interactive sessions.

%% ─────────────────────────────────────────────────────────────────────────
\section{The VM Memory Map}
\label{sec:memory-map}

StarForth allocates 5\,MB of virtual memory for the VM:

\begin{center}
\begin{tabular}{lll}
\toprule
Region & LBN range & Size \\
\midrule
Dictionary (word definitions) & LBN 0--2047 & 2\,MB \\
User blocks & LBN 2048 and above & user data \\
Log blocks & LBN 3072--5120 & 2\,MB \\
\bottomrule
\end{tabular}
\end{center}

\texttt{init.4th} is loaded from LBN 2048 at boot in the bare-metal kernel
configuration. On hosted Linux, \texttt{init.4th} is loaded from the
filesystem.

%% ─────────────────────────────────────────────────────────────────────────
\section{Block Words}
\label{sec:block-words}

\subsection*{BLOCK \texttt{( lbn -- addr )}}

Finds or loads block \texttt{lbn} into a buffer and pushes the buffer address.
The block is now in memory and can be read with \texttt{@} / \texttt{C@} or
written with \texttt{!} / \texttt{C!}.

\subsection*{BUFFER \texttt{( lbn -- addr )}}

Like \texttt{BLOCK} but does not read from storage: it claims a buffer for
\texttt{lbn} and leaves its contents uninitialized. Use when you intend to
write the entire block and do not need to read it first.

\subsection*{LOAD \texttt{( lbn -- )}}

Loads and executes the FORTH source in block \texttt{lbn}. This is the
traditional FORTH way to run stored code.

\subsection*{THRU \texttt{( lbn1 lbn2 -- )}}

Loads and executes blocks \texttt{lbn1} through \texttt{lbn2} inclusive.

\subsection*{FLUSH \texttt{( -- )}}

Writes all dirty buffers back to storage. Use before exiting or before
switching tasks if you have written to blocks.

\subsection*{UPDATE \texttt{( -- )}}

Marks the most recently referenced block buffer as dirty, so \texttt{FLUSH}
will write it back.

%%────────────────────────────────────────────────────────────────────────────
\chapter{String and I/O Words}
\label{chap:strings-io}
%%────────────────────────────────────────────────────────────────────────────

%% ─────────────────────────────────────────────────────────────────────────
\section{Output Words}
\label{sec:output}

\subsection*{EMIT \texttt{( char -- )}}

Outputs one character (given as an ASCII code).

\begin{lstlisting}[language=Forth]
ok> 65 EMIT
A ok>
\end{lstlisting}

\subsection*{CR \texttt{( -- )}}

Outputs a newline (carriage return and line feed).

\subsection*{SPACE \texttt{( -- )}}

Outputs a single space.

\subsection*{TYPE \texttt{( addr count -- )}}

Outputs \texttt{count} characters starting at \texttt{addr}.

\subsection*{.( \texttt{( "text" -- )}}

Immediately prints the text that follows (up to the closing parenthesis).
Works in both interpretation and compilation mode.

\begin{lstlisting}[language=Forth]
ok> .( Hello, world!)
Hello, world! ok>
\end{lstlisting}

\subsection*{." \texttt{( "text" -- )}}

Inside a word definition, compiles a string literal that prints when the word
executes:

\begin{lstlisting}[language=Forth]
: GREET  ." Hello from StarForth!" CR ;
\end{lstlisting}

%% ─────────────────────────────────────────────────────────────────────────
\section{Input Words}
\label{sec:input}

\subsection*{KEY \texttt{( -- char )}}

Waits for a keypress and pushes the ASCII code.

\subsection*{ACCEPT \texttt{( addr maxcount -- count )}}

Reads a line of input into the buffer at \texttt{addr}, up to
\texttt{maxcount} characters. Returns the actual count of characters read.

%% ─────────────────────────────────────────────────────────────────────────
\section{String Literals: S"}
\label{sec:s-quote}

\texttt{S"} compiles a string. At runtime it pushes the address and count of
the string data:

\begin{lstlisting}[language=Forth]
: PRINT-NAME  S" StarForth" TYPE CR ;
\end{lstlisting}

\begin{lstlisting}[language=Forth]
ok> PRINT-NAME
StarForth
ok>
\end{lstlisting}

%%────────────────────────────────────────────────────────────────────────────
\chapter{Running Files and Blocks}
\label{chap:files}
%%────────────────────────────────────────────────────────────────────────────

%% ─────────────────────────────────────────────────────────────────────────
\section{INCLUDE}
\label{sec:include}

\texttt{INCLUDE} loads a FORTH source file from the filesystem and executes
it:

\begin{lstlisting}[language=Forth]
ok> S" myprogram.4th" INCLUDE
ok>
\end{lstlisting}

All words defined in the file are added to the dictionary. Errors in the
included file are reported with the filename and line number.

%% ─────────────────────────────────────────────────────────────────────────
\section{EXEC}
\label{sec:exec}

\texttt{EXEC} loads a named capsule file (the StarForth extension for loading
\texttt{.4th} capsule files by name):

\begin{lstlisting}[language=Forth]
S" capsule.4th" EXEC
\end{lstlisting}

%% ─────────────────────────────────────────────────────────────────────────
\section{LOAD and THRU}
\label{sec:load-thru}

\texttt{LOAD} and \texttt{THRU} execute source from block storage (see
Chapter~\ref{chap:blocks}).

%%────────────────────────────────────────────────────────────────────────────
\chapter{The Interactive REPL}
\label{chap:repl}
%%────────────────────────────────────────────────────────────────────────────

%% ─────────────────────────────────────────────────────────────────────────
\section{The ok> Prompt}
\label{sec:ok-prompt}

The \texttt{ok>} prompt signals that StarForth is in interpretation mode,
the data stack is not empty but no error has occurred, and the system is
waiting for input. More precisely:

\begin{itemize}
  \item \texttt{ok>} — normal prompt; data stack may or may not be empty.
  \item After an error — StarForth prints the error, clears the stack to the
    pre-error state, and re-prints \texttt{ok>}.
\end{itemize}

%% ─────────────────────────────────────────────────────────────────────────
\section{Printing Numbers}
\label{sec:printing}

The basic output words:

\begin{center}
\begin{tabular}{lll}
\toprule
Word & Stack effect & Effect \\
\midrule
\texttt{.}  & \texttt{( n -- )} & Print top of stack in current base, then space \\
\texttt{.S} & \texttt{( -- )} & Print entire stack without consuming it \\
\texttt{.R} & \texttt{( n width -- )} & Print right-justified in field of width \\
\bottomrule
\end{tabular}
\end{center}

%% ─────────────────────────────────────────────────────────────────────────
\section{Numeric Base}
\label{sec:numeric-base}

StarForth defaults to decimal (base 10). Switch with:

\begin{lstlisting}[language=Forth]
ok> HEX
ok> FF .
255 ok>
ok> DECIMAL
ok>
\end{lstlisting}

\texttt{BASE} is a variable that holds the current numeric base (2--36).
\texttt{HEX} stores 16 in \texttt{BASE}; \texttt{DECIMAL} stores 10.

%% ─────────────────────────────────────────────────────────────────────────
\section{Dictionary Introspection}
\label{sec:introspection}

\begin{center}
\begin{tabular}{lll}
\toprule
Word & Stack effect & Description \\
\midrule
\texttt{WORDS}  & \texttt{( -- )} & Print all dictionary words \\
\texttt{FIND}   & \texttt{( addr len -- xt flags | 0 )} & Search by name \\
\texttt{'}      & \texttt{( "name" -- xt )} & Push execution token of named word \\
\texttt{SEE}    & \texttt{( "name" -- )} & Decompile a word \\
\bottomrule
\end{tabular}
\end{center}

%%━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━
\part{Physics Features}
%%━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━

\chapter{The Adaptive Runtime}
\label{chap:physics}

%% ─────────────────────────────────────────────────────────────────────────
\section{What the Adaptive Runtime Does for You}
\label{sec:physics-what}

While you type and run FORTH code, StarForth's adaptive runtime is watching in
the background. It tracks which words you execute most often, maintains a
rolling history of execution sequences, and adjusts internal caches and decay
rates to keep hot words fast and cold words from consuming resources.

From a user's perspective, you do not need to configure any of this. The
system tunes itself. The physics features become relevant when you want to
observe what the runtime is doing (\texttt{WORD-ENTROPY}), when you want to
lock behavior in place for reproducible benchmarks
(\texttt{PHYSICS-FREEZE}/\texttt{PHYSICS-THAW}), or when you run the Design
of Experiments mode (\texttt{--doe}).

The runtime uses thermodynamic language as a modeling convenience: ``heat'' is
execution frequency, ``decay'' is the process of cooling quiescent words,
``stability'' is a statistical property of the heat trajectory. These are
descriptive terms for internal metrics, not claims about physical
thermodynamics.

%% ─────────────────────────────────────────────────────────────────────────
\section{The Seven Feedback Loops}
\label{sec:seven-loops}

The adaptive runtime is governed by seven feedback loops running concurrently.
Here is a brief description of what each one does. For full technical detail,
see Volume~I of the StarForth VM and Physics Runtime reference.

\input{../scraps/architecture/03-architecture/physics-engine/feedback-loops}

%% ─────────────────────────────────────────────────────────────────────────
\section{WORD-ENTROPY}
\label{sec:word-entropy}

\texttt{WORD-ENTROPY} prints execution heat statistics for every word in the
dictionary to stdout. Use it to see which words are running hot:

\begin{lstlisting}[language=Forth]
ok> WORD-ENTROPY
\end{lstlisting}

The output lists each word with its current execution heat (a \Qtype{} value),
the number of times it has been called, and its position in the hot-words
cache (if it is cached). Words with zero heat have not been called since the
last reset.

\texttt{WORD-ENTROPY} is intentionally kept active in DoE mode because the
diagnostic value outweighs the output noise.

%% ─────────────────────────────────────────────────────────────────────────
\section{PHYSICS-FREEZE and PHYSICS-THAW}
\label{sec:freeze-thaw}

\texttt{PHYSICS-FREEZE} suspends all adaptive runtime updates. Heat values stop
changing, the decay loop stops, and the hot-words cache is locked in its
current state. Everything else --- word execution, I/O, compilation --- continues
normally.

\texttt{PHYSICS-THAW} resumes adaptive updates from the point they were frozen.

Use these words when you want reproducible timing:

\begin{lstlisting}[language=Forth]
ok> PHYSICS-FREEZE
ok> \ ... run your benchmark here ...
ok> PHYSICS-THAW
\end{lstlisting}

You might also freeze before a sensitive section to prevent cache eviction
from perturbing timing, then thaw afterward.

%% ─────────────────────────────────────────────────────────────────────────
\section{PHYSICS-RESET-STATS}
\label{sec:physics-reset}

\texttt{PHYSICS-RESET-STATS} clears all heat counters back to zero. Useful
before a fresh measurement run:

\begin{lstlisting}[language=Forth]
ok> PHYSICS-RESET-STATS
ok> \ ... warm up and then measure ...
\end{lstlisting}

%%────────────────────────────────────────────────────────────────────────────
\chapter{Heartbeat and Adaptive Tuning}
\label{chap:heartbeat}
%%────────────────────────────────────────────────────────────────────────────

\input{../scraps/architecture/heartbeat_csv_export}

%%────────────────────────────────────────────────────────────────────────────
\chapter{Design of Experiments Mode}
\label{chap:doe}
%%────────────────────────────────────────────────────────────────────────────

%% ─────────────────────────────────────────────────────────────────────────
\section{The --doe Flag}
\label{sec:doe-flag}

The \texttt{--doe} flag runs the full test harness in DoE (Design of
Experiments) mode and then exits:

\begin{lstlisting}[language=bash]
./build/amd64/fastest/starforth --doe
\end{lstlisting}

DoE mode exercises all 936+ tests and all seven feedback loops. It is used for
systematic experimental runs comparing different build configurations.

%% ─────────────────────────────────────────────────────────────────────────
\section{Capturing Heartbeat Metrics}
\label{sec:doe-capture}

In DoE mode, heartbeat metrics stream to \texttt{stderr}. Capture them for
analysis:

\begin{lstlisting}[language=bash]
./build/amd64/fastest/starforth --doe 2>heartbeat.csv 1>/dev/null
\end{lstlisting}

See Chapter~\ref{chap:heartbeat} for the CSV format and analysis guidance.

%% ─────────────────────────────────────────────────────────────────────────
\section{The 2\textsuperscript{7} Factorial Design}
\label{sec:doe-factorial}

DoE uses a 2\textsuperscript{7} factorial design, one factor per feedback loop.
Each run enables a specific combination of loops; the metrics measure the
incremental effect of each loop addition. The reference (baseline) has all
seven loops disabled. At the opposite extreme, all seven loops active, the
system is fully adaptive.

%% ─────────────────────────────────────────────────────────────────────────
\section{The Built-In Profiler}
\label{sec:profiler}

\input{../scraps/hardware/performance-profiling/PROFILER}

%%━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━
\part{The ACL System}
%%━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━

\chapter{Access Control for Interactive Users}
\label{chap:acl}

%% ─────────────────────────────────────────────────────────────────────────
\section{What ACL Means for You}
\label{sec:acl-user}

StarForth has a word-level access-control system. In a default development
build the ACL system is disabled (security is opt-in), so you will not
encounter it in most sessions.

When the ACL system is active, individual dictionary words may be restricted.
Attempting to execute a restricted word produces an error. Which words are
restricted, and under what conditions, is determined by the ACL policy loaded
at boot.

Most users on a standard build will never see ACL restrictions. The feature
is intended for deployments where StarForth serves as a controlled execution
environment (embedded systems, multi-user shells, kernel contexts).

%% ─────────────────────────────────────────────────────────────────────────
\section{The ok> Prompt vs.\ Emergency ok>}
\label{sec:acl-prompts}

There are two permanent console layers in StarForth:

\paragraph{Normal REPL --- \texttt{ok>}}
The standard interactive prompt. ACL restrictions are in effect if the ACL
system is active.

\paragraph{Emergency console --- \texttt{ok>}}
When a fault handler activates (a crash, an unrecoverable error), StarForth
falls back to the emergency console. The emergency console has a 100\% ACL
bypass --- all words execute regardless of their ACL state. This ensures you
can always recover, even in a locked-down deployment.

The \texttt{EMERGENCY\_CONSOLE\_ENABLED} build flag (default: on) controls
whether the emergency console is present. In production embedded builds it may
be disabled.

%% ─────────────────────────────────────────────────────────────────────────
\section{The Zuse Session}
\label{sec:acl-zuse}

\texttt{zuse} is the bootstrap superuser in StarForth's security model. When a
Zuse session is active, the prompt changes to:

\begin{lstlisting}[language=Forth]
zuse)ok>
\end{lstlisting}

A Zuse session has a full ACL bypass: all words execute regardless of their
normal ACL state. The Zuse console is completely independent of the emergency
console --- either may be active without affecting the other.

How to authenticate as Zuse and what operations a Zuse session enables is
covered in the StarForth Developer Guide, not here. From a user's perspective,
if you see \texttt{zuse)ok>}, you are in a privileged session.

%% ─────────────────────────────────────────────────────────────────────────
\section{Enabling the ACL System}
\label{sec:acl-enable}

The ACL system is opt-in. In \texttt{init.4th}, one commented-out line
controls it:

\begin{lstlisting}[language=Forth]
\ S" ACL.4th" EXEC
\end{lstlisting}

Uncommenting that line enables full ACL enforcement at boot. Leaving it
commented keeps a development build open. See the StarForth Developer Guide
for the full ACL design.

%%────────────────────────────────────────────────────────────────────────────
%% FORTH Programming — patterns, debugging, working with numbers, vocabularies,
%% I/O, source files, tests
%%────────────────────────────────────────────────────────────────────────────
\input{ch-forth-programming}

%%────────────────────────────────────────────────────────────────────────────
%% Advanced Topics — VM internals, performance tuning, block editor, capsules,
%% Jacquard, formal verification, embedded targets
%%────────────────────────────────────────────────────────────────────────────
\input{ch-advanced}

%%────────────────────────────────────────────────────────────────────────────
%% Recipes — idioms, worked examples, migration guide
%%────────────────────────────────────────────────────────────────────────────
\input{ch-recipes}

%%━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━
\part{Running on Hardware}
%%━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━

\chapter{Linux and POSIX Platforms}
\label{chap:linux}

StarForth's standard build requires only GCC (or Clang) and GNU Make. No
additional runtime dependencies are needed beyond the standard C library.

\begin{lstlisting}[language=bash]
make fastest
./build/amd64/standard/starforth
\end{lstlisting}

The binary is statically linked on supported configurations. No shared
libraries are needed on the target system.

%% TODO(bob): add notes on L4Re microkernel integration once status is confirmed.

\chapter{Raspberry Pi 4}
\input{../scraps/hardware/platform-integration/RASPBERRY_PI_BUILD}

\chapter{LithosAnanke Bare-Metal Boot}
\label{chap:lithos}

LithosAnanke (``stone inevitability'') is the bare-metal UEFI kernel that boots
StarForth directly on hardware without a host operating system. Version 1.5.3
ships as a monolithic UEFI PE32+ executable.

%% ─────────────────────────────────────────────────────────────────────────
\section{Boot Sequence}
\label{sec:lithos-boot}

\begin{lstlisting}[language=bash]
UEFI Firmware -> uefi_loader.c (BOOTX64.EFI)
    ExitBootServices() -- kernel owns hardware
    -> kernel_main()
        M1: Console (UART 16550 + framebuffer)
        M2: Physical memory manager
        M3: Virtual memory (4-level paging)
        M4: IDT + APIC interrupts
        M5: TSC + HPET + APIC timer (100 Hz heartbeat)
        M6: kmalloc heap
        M7: StarForth VM + capsule loading -> ok> REPL
\end{lstlisting}

Milestones M0--M6 are complete. M7 (StarForth VM integration) is in progress.

%% ─────────────────────────────────────────────────────────────────────────
\section{Building the Kernel}
\label{sec:lithos-build}

\begin{lstlisting}[language=bash]
# AMD64
make -f Makefile.starkernel

# AArch64
make -f Makefile.starkernel ARCH=aarch64

# RISC-V 64
make -f Makefile.starkernel ARCH=riscv64

# Run in QEMU with OVMF
make -f Makefile.starkernel qemu
\end{lstlisting}

Output: \texttt{build/amd64/kernel/starkernel\_loader.efi} (UEFI PE32+ executable).

%% TODO(bob): add hardware bring-up notes once M7 acceptance testing is complete.

%%━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━
%% Appendices
%%━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━
\appendix

\chapter{Build Flags Reference}
\input{../scraps/architecture/build-and-tooling/BUILD_OPTIONS}

%%────────────────────────────────────────────────────────────────────────────
\chapter{FORTH-79 Word Quick Reference}
\label{app:word-ref}
%%────────────────────────────────────────────────────────────────────────────

This table covers the full FORTH-79 word set available in StarForth, organized
alphabetically within categories. Stack effects use standard notation:
\texttt{( before -- after )}, where items are listed bottom-to-top and the
rightmost item is the top of stack.

Flags are \texttt{-1} (true) or \texttt{0} (false) unless noted otherwise.
Addresses (\texttt{addr}) are VM byte offsets, not C pointers.

%% ─────────────────────────────────────────────────────────────────────────
\section*{Stack Words}
\begin{longtable}{>{\ttfamily}lp{4.5cm}p{7cm}}
\toprule
\normalfont Word & Stack Effect & Description \\
\midrule
\endhead
\bottomrule
\endfoot
DUP     & ( n -- n n )        & Duplicate top item \\
DROP    & ( n -- )            & Discard top item \\
SWAP    & ( a b -- b a )      & Exchange top two items \\
ROT     & ( a b c -- b c a )  & Rotate top three; third comes to top \\
OVER    & ( a b -- a b a )    & Copy second item to top \\
NIP     & ( a b -- b )        & Drop second item \\
TUCK    & ( a b -- b a b )    & Copy top item below second \\
2DUP    & ( a b -- a b a b )  & Duplicate top pair \\
2DROP   & ( a b -- )          & Drop top pair \\
2SWAP   & ( a b c d -- c d a b ) & Exchange top two pairs \\
2OVER   & ( a b c d -- a b c d a b ) & Copy second pair to top \\
DEPTH   & ( -- n )            & Number of items on the data stack \\
PICK    & ( xn...x0 n -- xn...x0 xn ) & Copy nth item to top \\
ROLL    & ( xn...x0 n -- xn-1...x0 xn ) & Rotate nth item to top \\
\end{longtable}

%% ─────────────────────────────────────────────────────────────────────────
\section*{Return Stack Words}
\begin{longtable}{>{\ttfamily}lp{4.5cm}p{7cm}}
\toprule
\normalfont Word & Stack Effect & Description \\
\midrule
\endhead
\bottomrule
\endfoot
>R      & ( n -- ) (R: -- n )        & Move top data item to return stack \\
R>      & ( -- n ) (R: n -- )        & Move top return item to data stack \\
R@      & ( -- n ) (R: n -- n )      & Copy top of return stack to data stack \\
RDROP   & ( -- ) (R: n -- )          & Discard top of return stack \\
2>R     & ( a b -- ) (R: -- a b )    & Move top data pair to return stack \\
2R>     & ( -- a b ) (R: a b -- )    & Move top return pair to data stack \\
2R@     & ( -- a b ) (R: a b -- a b ) & Copy top return pair to data stack \\
\end{longtable}

%% ─────────────────────────────────────────────────────────────────────────
\section*{Arithmetic Words}
\begin{longtable}{>{\ttfamily}lp{4.5cm}p{7cm}}
\toprule
\normalfont Word & Stack Effect & Description \\
\midrule
\endhead
\bottomrule
\endfoot
+       & ( a b -- a+b )      & Integer addition \\
-       & ( a b -- a-b )      & Integer subtraction \\
*       & ( a b -- a*b )      & Integer multiplication \\
/       & ( a b -- a/b )      & Integer division (truncate toward zero) \\
MOD     & ( a b -- a mod b )  & Integer remainder \\
/MOD    & ( a b -- rem quot ) & Division and remainder \\
ABS     & ( n -- |n| )        & Absolute value \\
NEGATE  & ( n -- -n )         & Arithmetic negation \\
MIN     & ( a b -- min )      & Minimum of two values \\
MAX     & ( a b -- max )      & Maximum of two values \\
1+      & ( n -- n+1 )        & Add 1 \\
1-      & ( n -- n-1 )        & Subtract 1 \\
2+      & ( n -- n+2 )        & Add 2 \\
2-      & ( n -- n-2 )        & Subtract 2 \\
2*      & ( n -- n*2 )        & Multiply by 2 (left shift 1) \\
2/      & ( n -- n/2 )        & Divide by 2 (arithmetic right shift 1) \\
\end{longtable}

%% ─────────────────────────────────────────────────────────────────────────
\section*{Double-Cell Arithmetic}
\begin{longtable}{>{\ttfamily}lp{4.5cm}p{7cm}}
\toprule
\normalfont Word & Stack Effect & Description \\
\midrule
\endhead
\bottomrule
\endfoot
D+      & ( d1 d2 -- d1+d2 )  & Double-cell addition \\
D-      & ( d1 d2 -- d1-d2 )  & Double-cell subtraction \\
D=      & ( d1 d2 -- flag )    & Double-cell equality \\
D<      & ( d1 d2 -- flag )    & Double-cell signed less-than \\
DNEGATE & ( d -- -d )          & Double-cell negation \\
DABS    & ( d -- |d| )         & Double-cell absolute value \\
M+      & ( d n -- d+n )       & Add single to double \\
M*      & ( a b -- d )         & Multiply; produce double-cell result \\
UM*     & ( u1 u2 -- ud )      & Unsigned multiply; produce double-cell result \\
\end{longtable}

%% ─────────────────────────────────────────────────────────────────────────
\section*{Logical and Bitwise Words}
\begin{longtable}{>{\ttfamily}lp{4.5cm}p{7cm}}
\toprule
\normalfont Word & Stack Effect & Description \\
\midrule
\endhead
\bottomrule
\endfoot
AND     & ( a b -- a\&b )     & Bitwise AND (also logical AND on flags) \\
OR      & ( a b -- a|b )      & Bitwise OR (also logical OR on flags) \\
XOR     & ( a b -- a xor b )  & Bitwise exclusive OR \\
INVERT  & ( n -- \textasciitilde n ) & Bitwise complement (all bits flipped) \\
NOT     & ( n -- flag )       & Logical NOT: -1 if n=0, else 0 \\
LSHIFT  & ( n count -- n' )   & Left shift n by count bits \\
RSHIFT  & ( n count -- n' )   & Logical right shift n by count bits \\
\end{longtable}

%% ─────────────────────────────────────────────────────────────────────────
\section*{Comparison Words}
\begin{longtable}{>{\ttfamily}lp{4.5cm}p{7cm}}
\toprule
\normalfont Word & Stack Effect & Description \\
\midrule
\endhead
\bottomrule
\endfoot
=       & ( a b -- flag )     & Equal \\
<>      & ( a b -- flag )     & Not equal \\
<       & ( a b -- flag )     & Signed less-than \\
>       & ( a b -- flag )     & Signed greater-than \\
<=      & ( a b -- flag )     & Signed less-than-or-equal \\
>=      & ( a b -- flag )     & Signed greater-than-or-equal \\
U<      & ( u1 u2 -- flag )   & Unsigned less-than \\
U>      & ( u1 u2 -- flag )   & Unsigned greater-than \\
0=      & ( n -- flag )       & True if n equals zero \\
0<>     & ( n -- flag )       & True if n is non-zero \\
0<      & ( n -- flag )       & True if n is negative \\
0>      & ( n -- flag )       & True if n is positive \\
\end{longtable}

%% ─────────────────────────────────────────────────────────────────────────
\section*{Memory Words}
\begin{longtable}{>{\ttfamily}lp{4.5cm}p{7cm}}
\toprule
\normalfont Word & Stack Effect & Description \\
\midrule
\endhead
\bottomrule
\endfoot
@       & ( addr -- n )       & Fetch one cell from addr \\
!       & ( n addr -- )       & Store one cell at addr \\
C@      & ( addr -- char )    & Fetch one byte from addr \\
C!      & ( char addr -- )    & Store one byte at addr \\
2@      & ( addr -- d )       & Fetch double-cell (two consecutive cells) \\
2!      & ( d addr -- )       & Store double-cell \\
MOVE    & ( src dst count -- ) & Copy count bytes from src to dst \\
FILL    & ( addr count char -- ) & Fill count bytes at addr with char \\
HERE    & ( -- addr )         & Address of next free dictionary byte \\
ALLOT   & ( n -- )            & Reserve n bytes in dictionary \\
,       & ( n -- )            & Compile n as a cell into dictionary \\
C,      & ( char -- )         & Compile char as a byte into dictionary \\
CELLS   & ( n -- n*cell-size ) & Convert cell count to byte count \\
CELL+   & ( addr -- addr' )   & Advance address by one cell \\
CHARS   & ( n -- n )          & Convert char count to byte count (=1 on byte-addressed) \\
\end{longtable}

%% ─────────────────────────────────────────────────────────────────────────
\section*{Control Flow Words}
\begin{longtable}{>{\ttfamily}lp{4.5cm}p{7cm}}
\toprule
\normalfont Word & Stack Effect & Description \\
\midrule
\endhead
\bottomrule
\endfoot
IF      & ( flag -- )         & Branch if flag is zero (skip to ELSE/THEN) \\
ELSE    & ( -- )              & Alternate branch after IF \\
THEN    & ( -- )              & End of IF...ELSE...THEN \\
DO      & ( limit start -- )  & Begin counted loop \\
LOOP    & ( -- )              & Increment index by 1; exit if index = limit \\
+LOOP   & ( n -- )            & Increment index by n; exit when index crosses limit \\
LEAVE   & ( -- )              & Exit enclosing DO loop immediately \\
I       & ( -- index )        & Copy innermost loop index \\
J       & ( -- index )        & Copy outermost loop index \\
BEGIN   & ( -- )              & Mark start of BEGIN loop \\
UNTIL   & ( flag -- )         & Branch back to BEGIN if flag is zero \\
WHILE   & ( flag -- )         & Exit BEGIN loop if flag is zero \\
REPEAT  & ( -- )              & Branch back to BEGIN (used with WHILE) \\
AGAIN   & ( -- )              & Unconditional branch back to BEGIN \\
EXIT    & ( -- )              & Return from current word \\
EXECUTE & ( xt -- )           & Execute word whose execution token is xt \\
\end{longtable}

%% ─────────────────────────────────────────────────────────────────────────
\section*{Defining Words}
\begin{longtable}{>{\ttfamily}lp{4.5cm}p{7cm}}
\toprule
\normalfont Word & Stack Effect & Description \\
\midrule
\endhead
\bottomrule
\endfoot
:       & ( "name" -- )       & Begin word definition; enter compilation mode \\
;       & ( -- )              & End word definition; return to interpretation mode \\
IMMEDIATE & ( -- )            & Mark most recently defined word as immediate \\
CONSTANT & ( n "name" -- )   & Define a constant word that pushes n \\
VARIABLE & ( "name" -- )     & Define a variable; word pushes its address \\
CREATE  & ( "name" -- )      & Create a named dictionary entry; no behavior \\
DOES>   & ( -- )             & Specify runtime behavior for CREATE'd words \\
FORGET  & ( "name" -- )      & Remove named word and all newer words from dictionary \\
'       & ( "name" -- xt )   & Push execution token of named word \\
{[}'{}]     & ( "name" -- xt )   & Compile-time version of tick (immediate) \\
\end{longtable}

%% ─────────────────────────────────────────────────────────────────────────
\section*{I/O and String Words}
\begin{longtable}{>{\ttfamily}lp{4.5cm}p{7cm}}
\toprule
\normalfont Word & Stack Effect & Description \\
\midrule
\endhead
\bottomrule
\endfoot
EMIT    & ( char -- )         & Output one character \\
KEY     & ( -- char )         & Wait for and return one keypress \\
CR      & ( -- )              & Output newline \\
SPACE   & ( -- )              & Output one space \\
SPACES  & ( n -- )            & Output n spaces \\
TYPE    & ( addr count -- )   & Output count characters from addr \\
.       & ( n -- )            & Print top of stack in current base, followed by space \\
.S      & ( -- )              & Print entire stack contents without consuming \\
.R      & ( n width -- )      & Print n right-justified in field of width \\
U.      & ( u -- )            & Print top of stack as unsigned \\
U.R     & ( u width -- )      & Print unsigned right-justified in field \\
.(      & ( "text)" -- )      & Immediately print text up to closing paren \\
."      & ( "text" -- )       & Compile string that prints at runtime \\
S"      & ( "text" -- addr count ) & Compile string literal; push addr+count at runtime \\
ACCEPT  & ( addr maxcount -- count ) & Read line of input into buffer \\
WORD    & ( char -- addr )    & Parse next word from input, delimited by char \\
PARSE   & ( char -- addr count ) & Parse input to delimiter, return addr+count \\
\end{longtable}

%% ─────────────────────────────────────────────────────────────────────────
\section*{Numeric Output Format Words}
\begin{longtable}{>{\ttfamily}lp{4.5cm}p{7cm}}
\toprule
\normalfont Word & Stack Effect & Description \\
\midrule
\endhead
\bottomrule
\endfoot
HEX     & ( -- )              & Set numeric base to 16 \\
DECIMAL & ( -- )              & Set numeric base to 10 \\
BINARY  & ( -- )              & Set numeric base to 2 \\
BASE    & ( -- addr )         & Address of the base variable (default 10) \\
<\#     & ( -- )              & Begin pictured numeric output \\
\#      & ( ud -- ud' )       & Convert one digit of pictured numeric output \\
\#S     & ( ud -- 0 0 )       & Convert all remaining digits \\
HOLD    & ( char -- )         & Insert char into pictured numeric output \\
\#>     & ( xd -- addr count ) & End pictured numeric output; return string \\
SIGN    & ( n -- )            & Add minus sign to pictured output if n is negative \\
\end{longtable}

%% ─────────────────────────────────────────────────────────────────────────
\section*{Block Words}
\begin{longtable}{>{\ttfamily}lp{4.5cm}p{7cm}}
\toprule
\normalfont Word & Stack Effect & Description \\
\midrule
\endhead
\bottomrule
\endfoot
BLOCK   & ( lbn -- addr )     & Load block lbn into a buffer; return buffer address \\
BUFFER  & ( lbn -- addr )     & Assign buffer to lbn without reading; return address \\
LOAD    & ( lbn -- )          & Load and execute FORTH source in block lbn \\
THRU    & ( lbn1 lbn2 -- )    & Load and execute blocks lbn1 through lbn2 \\
FLUSH   & ( -- )              & Write all dirty buffers to storage \\
UPDATE  & ( -- )              & Mark current buffer as dirty (modified) \\
SAVE-BUFFERS & ( -- )         & Synonym for FLUSH \\
EMPTY-BUFFERS & ( -- )        & Invalidate all buffers without writing \\
\end{longtable}

%% ─────────────────────────────────────────────────────────────────────────
\section*{Dictionary and System Words}
\begin{longtable}{>{\ttfamily}lp{4.5cm}p{7cm}}
\toprule
\normalfont Word & Stack Effect & Description \\
\midrule
\endhead
\bottomrule
\endfoot
WORDS   & ( -- )              & Print all dictionary words \\
FIND    & ( addr len -- xt flags | 0 ) & Search dictionary by name \\
SEARCH-WORDLIST & ( addr len wid -- xt flags | 0 ) & Search a specific wordlist \\
VOCABULARY & ( "name" -- )   & Define a new vocabulary (wordlist) \\
DEFINITIONS & ( -- )         & Set current vocabulary as definition target \\
FORTH-WORDLIST & ( -- wid )  & Push the id of the FORTH vocabulary \\
STATE   & ( -- addr )         & Address of compilation-state flag (0=interpret, -1=compile) \\
SEE     & ( "name" -- )       & Decompile named word \\
ABORT   & ( -- )              & Reset stacks, print error message, restart interpreter \\
ABORT"  & ( flag "text" -- )  & If flag is non-zero, abort with message \\
QUIT    & ( -- )              & Return to outer interpreter (no error message) \\
BYE     & ( -- )              & Exit StarForth \\
INCLUDE & ( "filename" -- )   & Load and execute FORTH source file \\
\end{longtable}

%% ─────────────────────────────────────────────────────────────────────────
\section*{Q48.16 Fixed-Point Words}
\begin{longtable}{>{\ttfamily}lp{4.5cm}p{7cm}}
\toprule
\normalfont Word & Stack Effect & Description \\
\midrule
\endhead
\bottomrule
\endfoot
Q+      & ( q1 q2 -- q1+q2 )  & \Qtype{} addition \\
Q-      & ( q1 q2 -- q1-q2 )  & \Qtype{} subtraction \\
Q*      & ( q1 q2 -- q1*q2 )  & \Qtype{} multiplication \\
Q/      & ( q1 q2 -- q1/q2 )  & \Qtype{} division \\
INT>Q   & ( n -- q )           & Convert integer to \Qtype{} \\
Q>INT   & ( q -- n )           & Convert \Qtype{} to integer (truncate) \\
Q.      & ( q -- )             & Print \Qtype{} value as decimal fraction \\
\end{longtable}

%% ─────────────────────────────────────────────────────────────────────────
\section*{StarForth Physics Words}
\begin{longtable}{>{\ttfamily}lp{4.5cm}p{7cm}}
\toprule
\normalfont Word & Stack Effect & Description \\
\midrule
\endhead
\bottomrule
\endfoot
WORD-ENTROPY        & ( -- )   & Print execution heat statistics for all words \\
PHYSICS-FREEZE      & ( -- )   & Suspend all adaptive runtime updates \\
PHYSICS-THAW        & ( -- )   & Resume adaptive runtime updates after freeze \\
PHYSICS-RESET-STATS & ( -- )   & Clear all heat counters to zero \\
\end{longtable}

%% ─────────────────────────────────────────────────────────────────────────
\section*{ACL Words (active only when ACL.4th is loaded)}
\begin{longtable}{>{\ttfamily}lp{4.5cm}p{7cm}}
\toprule
\normalfont Word & Stack Effect & Description \\
\midrule
\endhead
\bottomrule
\endfoot
ACL-ENTRY       & ( xt -- addr )      & O(1) ACL table lookup by execution token \\
ACL-MODE@       & ( xt -- mode )      & Read enforcement mode of word \\
ACL-MODE!       & ( mode xt -- )      & Set enforcement mode (no-op if pinned) \\
ACL-PINNED?     & ( xt -- flag )      & Test whether word is pinned \\
ACL-PIN         & ( xt -- )           & One-way pin: freeze mode and decision permanently \\
ACL-STRICT      & ( xt -- )           & Set STRICT enforcement mode (no-op if pinned) \\
ACL-TTL-MODE    & ( xt -- )           & Set adaptive TTL mode (no-op if pinned) \\
ACL-TTL@        & ( xt -- n )         & Read current TTL counter \\
ACL-TTL!        & ( n xt -- )         & Write TTL counter \\
ACL-ALLOW@      & ( xt -- flag )      & Read cached access decision \\
ACL-ALLOW!      & ( flag xt -- )      & Write access decision \\
ACL-INHERIT     & ( src dst -- )      & Copy mode from src to dst; clear pin; reset TTL and decision \\
ACL-RECHECK     & ( xt -- )           & Recompute adaptive TTL and update decision \\
ACL-INIT-PRIMITIVES & ( -- )          & Bulk-initialize ACL entries for all existing words \\
ACL-GRANT       & ( xt -- )           & Grant access to word \\
ACL-REVOKE      & ( xt -- )           & Revoke access to word \\
\end{longtable}

%%────────────────────────────────────────────────────────────────────────────
\chapter{Glossary}
\label{app:glossary}
%%────────────────────────────────────────────────────────────────────────────

\begin{description}
  \item[ACL] Access Control List. In StarForth, a per-word access-control
    mechanism implemented in \texttt{ACL.4th}.
  \item[Capsule] An immutable, content-addressed VM initialization payload.
    A capsule's ID is its XXHash64 content hash.
  \item[Cell] The fundamental unit of the data stack. On a 64-bit host,
    one cell is 64 bits (8 bytes).
  \item[Colon definition] A word defined with \texttt{:} and \texttt{;}.
  \item[Compilation mode] The interpreter state entered after \texttt{:},
    in which words are compiled rather than executed.
  \item[Data stack] The primary working stack, distinct from the return stack.
  \item[Decay] The process by which execution heat diminishes over time for
    quiescent words (\loopref{3}).
  \item[DoE] Design of Experiments. StarForth's 2\textsuperscript{7} factorial
    experimental framework for measuring the effect of each feedback loop.
  \item[Execution token (XT)] An opaque value identifying a word in the
    dictionary, suitable for use with \texttt{EXECUTE} or ACL words.
  \item[Heat] The execution-frequency metric tracked per word by
    \loopref{1}. Higher heat means more recent and frequent execution.
  \item[Immediate word] A word that executes at compile time rather than
    being compiled into the current word definition.
  \item[Interpretation mode] The default interpreter state in which words
    are executed as they are encountered.
  \item[Jacquard Selector] The L8 steady-state machine (\texttt{ssm\_jacquard.c})
    that selects between runtime mode configurations: stable, volatile,
    diverse, temporal, transition, omni.
  \item[LBN] Logical Block Number. The address scheme for StarForth's
    block storage subsystem.
  \item[LithosAnanke] The bare-metal UEFI kernel that boots StarForth
    directly on hardware without a host operating system.
  \item[\Qtype{}] Q48.16 fixed-point format: 48-bit integer part, 16-bit
    fractional part. Precision of $1/65536 \approx 0.0000153$.
  \item[Return stack] The secondary stack used for word-call linkage and
    temporary storage with \texttt{>R}/\texttt{R>}.
  \item[Rolling window] The circular execution history buffer managed by
    \loopref{2}.
  \item[SMUDGE] A flag set during word compilation that hides the partial
    definition from dictionary searches until \texttt{;} completes it.
  \item[SSM] Steady-State Machine. The control structure governing
    StarForth's adaptive runtime.
  \item[Stack effect] The notation \texttt{( before -- after )} documenting
    what a word consumes and produces on the data stack.
  \item[Zuse] The bootstrap superuser in StarForth's security model, named
    for Konrad Zuse.
\end{description}

\backmatter
\bibliographystyle{plainnat}
\bibliography{../common/starship}

\end{document}
