%% SCRAP: archive/session-logs/README
%% SOURCE: docs/working/archive/session-logs/README.md
%% STATUS: HISTORICAL
%% FITS: none
%% EDITORIAL: lifted — prose rewritten to press voice

\section{Session Logs: Index and Policy}

This directory holds historical development session summaries preserved for
context and reference. The documents are informal notes from individual
working sessions; they are not maintained and may contain outdated
information.

\subsection{When Session Logs Are Created}

Session logs are appropriate for:
\begin{itemize}
  \item Multi-hour debugging sessions with significant discoveries.
  \item Experimental work involving multiple failed attempts before success.
  \item Design discussions that drove important architectural decisions.
  \item Complex refactoring with non-obvious rationale.
\end{itemize}

They are not appropriate substitutes for routine git commit messages,
formal architecture documentation, or final implementation records.

\subsection{Contents}

\begin{itemize}
  \item \texttt{2025-11-20-recap.md} — Paradigm shift from factorial DoE to
    multivariate coupled-dynamics analysis.
  \item \texttt{2025-11-26-worklog.md} — 2$^7$ DoE analysis completion and
    L8 Jacquard mode selector final integration.
  \item \texttt{experimental-addendum.md} — Discovery narrative for James Law
    of Computational Dynamics.
  \item \texttt{experimental-iteration-usage.md} — Experimental iteration
    runner usage guide.
\end{itemize}

\subsection{Relationship to Formal Documentation}

Session logs supplement formal documentation; they do not replace it. They
are point-in-time snapshots written from a developer perspective. Readers
seeking current architectural information should consult the formal volume.
