%% SCRAP: archive/quality/audits/section-2-audit
%% SOURCE: docs/working/archive/quality/audits/section-2-audit.md
%% STATUS: HISTORICAL
%% FITS: none
%% EDITORIAL: lifted — prose rewritten to press voice

\section*{Section~2 Audit: Physics Subsystem Safety \& Clarity (November 2025)}

A safety audit of five physics subsystem files conducted on 2025-11-20
classified the subsystem as broadly safe with no critical vulnerabilities.
Results by file:

\begin{itemize}
  \item \texttt{physics\_metadata.c}: All functions carry \texttt{NULL} checks.
        Bounds clamping prevents overflow and underflow on heat accumulation
        and decay paths. Compile-time guards were absent.
  \item \texttt{physics\_pipelining\_metrics.c}: Two high-priority issues
        identified---an unchecked array access in \texttt{transition\_metrics\_update\_cache()}
        when \texttt{dict\_size} changes between allocation and use, and a
        missing bounds check on \texttt{target\_word\_id} in
        \texttt{transition\_metrics\_get\_probability\_q48()}.
  \item \texttt{physics\_hotwords\_cache.c}: Comprehensive \texttt{NULL} and
        bounds checks; LRU round-robin eviction correctly capped.
  \item \texttt{rolling\_window\_of\_truth.c}: Assessed as best defensive
        coding of all files; modulo arithmetic, double-buffered atomic
        snapshots, and depth-clamping all correctly implemented.
  \item \texttt{physics\_runtime.c}: Not audited in this section; deferred
        as out-of-scope (host analytics heap, not word-dispatch logic).
\end{itemize}

Recommendations covered 15 intent-focused comment blocks (feedback-loop
annotations), 2 bounds-check additions, compile-time guard wrappers for 4
files, and a log-level downgrade to suppress \textsc{doe} mode stderr
pollution. All recommendations were addressed in the corresponding corrective
actions (see Section~2 Executive Summary in this archive).
