%% SCRAP: archive/legacy/phase-1/Reference/codex_audit_report
%% SOURCE: docs/working/archive/legacy/phase-1/Reference/codex_audit_report.md
%% STATUS: HISTORICAL
%% FITS: none
%% EDITORIAL: lifted — prose rewritten to press voice

\section{Governance Audit Report --- Phase 1 (2025)}

An independent audit of the \texttt{StarForth-Governance} repository identified
42~nonconformities spanning validation evidence, coverage metrics, platform testing
records, process signatures, and CI/CD artefact retention. The audit returned a verdict
of \textbf{not audit-ready}. All major release gates were unsatisfied at the time of
the audit.

\subsection{Finding Categories}

\begin{itemize}
  \item \textbf{Missing validation artefacts (F-01 to F-09, F-18).}
        Required documents absent: ANSI~C99 compliance matrix, complete FORTH-79 word
        compliance matrix (65 of 70 words missing), ARM64 test logs,
        embedded platform evidence, and the referenced HTML coverage report.
  \item \textbf{Placeholder data in signed records (F-07, F-10 to F-13, F-19, F-20).}
        Coverage metrics (\texttt{\_\_\%}), deficiency counts (\texttt{[N]}), and
        timing variance tables (\texttt{X}) remained unfilled in documents that had
        been marked complete or passed.
  \item \textbf{Missing signatures (F-15 to F-17, F-22 to F-34).}
        Validation engineer, review authority, and maintainer sign-offs absent from
        Tier~I and Tier~II final reports, all governance foundation documents, and all
        process chapters (ECR, ECO, CAPA, FMEA, theory justification guide).
  \item \textbf{Governance self-conflicts (F-35 to F-42).}
        Documents simultaneously asserted completion and contained unfilled placeholders;
        the governance model prescribed continuous validation CI evidence that did not
        exist; deficiency logs were cited as proof of zero deficiencies while containing
        only a blank template.
\end{itemize}

\subsection{Remediation Register}

The audit generated 42~CAPAs (CAPA-01 through CAPA-42) covering: authoring the missing
compliance matrices, executing and archiving multi-platform tests, populating all
placeholder fields, securing signatures across all policy and process documents,
implementing continuous validation CI evidence export, and standing up static analysis
(SonarQube or equivalent) to satisfy ISO~5055 quality gate requirements.

%% TODO(bob): record the current remediation status --- which CAPAs were closed before
%% the governance effort was paused or superseded.
