%% SCRAP: archive/legacy/phase-1/Reference/physics_experiment/PEER_REVIEW_SUBMISSION/05_SUPPLEMENTARY/PEER_REVIEW_PAPER_OUTLINE
%% SOURCE: docs/working/archive/legacy/phase-1/Reference/physics_experiment/PEER_REVIEW_SUBMISSION/05_SUPPLEMENTARY/PEER_REVIEW_PAPER_OUTLINE.md
%% STATUS: SUPERSEDED
%% FITS: none
%% EDITORIAL: lifted — prose rewritten to press voice

\section{Paper Outline: Physics-Driven Self-Adaptive FORTH-79 VM}

This outline was prepared as a submission scaffold for the peer review paper
subsequently published on SSRN (abstract 5930134). It is preserved as a
historical record of the paper's structural evolution. The published paper
supersedes all content below.

\subsection{Abstract (archived draft)}

The paper presented a formally verified self-adaptive FORTH-79 virtual machine
employing a physics-based runtime model to dynamically optimize execution
configuration. The system tracked \texttt{execution\_heat} as a
frequency-based optimization metric and converged toward near-optimal
performance through selective hotword caching and predictive prefetching.

Key contributions claimed in the draft:
\begin{itemize}
  \item Physics model enables deterministic, observable optimization
        (cache hit rate variance: 0\%).
  \item Measurable convergence: 25.4\% performance improvement in the
        adaptive configuration.
  \item Formal stability: algorithmic determinism maintained despite 70\%
        environmental variance.
  \item Practical performance bounds established via a 90-run validation
        campaign.
\end{itemize}

\subsection{Experimental Methodology (archived draft)}

The draft described a 90-run randomized experiment (30 runs per
configuration) across three configurations:

\begin{center}
\begin{tabular}{lllp{5cm}}
\toprule
Configuration & Cache & Pipelining & Purpose \\
\midrule
A\_BASELINE & off & off & Control (no optimization) \\
B\_CACHE    & on  & off & Baseline adaptive \\
C\_FULL     & on  & on  & Full physics model \\
\bottomrule
\end{tabular}
\end{center}

33 metrics were collected per run, including runtime, cache hit/miss counts,
latency, and CPU thermal state.

\subsection{Formal Claims (archived draft)}

The draft enumerated four theorems:

\begin{description}
  \item[Theorem 1 — Algorithmic Determinism.]
        Cache hit rates exhibited 0\% coefficient of variation across all
        30 runs per configuration, demonstrating deterministic state
        transitions.
  \item[Theorem 2 — Adaptive Convergence.]
        C\_FULL runtime decreased from 10.20\,ms (early runs 1--15) to
        7.61\,ms (late runs 16--30), a 25.4\% monotonic improvement.
  \item[Theorem 3 — Environmental Robustness.]
        OS scheduling produced 70\% runtime CV; algorithmic CV remained 0\%,
        establishing complete independence of internal state from external
        noise.
  \item[Theorem 4 — Predictable Performance Bounds.]
        95\% confidence intervals established: B\_CACHE $7.80 \pm 5.04$\,ms.
\end{description}

%% TODO(bob): cite final SSRN publication (abstract 5930134) wherever
%% this outline is referenced in the formal volumes.
