%% SCRAP: archive/legacy/phase-1/Reference/physics_experiment/PEER_REVIEW_SUBMISSION/05_SUPPLEMENTARY/WORK_COMPLETION_STATUS
%% SOURCE: docs/working/archive/legacy/phase-1/Reference/physics_experiment/PEER_REVIEW_SUBMISSION/05_SUPPLEMENTARY/WORK_COMPLETION_STATUS.md
%% STATUS: HISTORICAL
%% FITS: none
%% EDITORIAL: lifted — prose rewritten to press voice

\section{Pre-Submission Work Completion Status (November 2025)}

This status report, dated 7 November 2025, captures the state of the
physics-driven FORTH-79 VM paper immediately prior to its final publication
push. The experimental and analytical foundation was complete; remaining
work was preparation and formatting. The paper was subsequently published
on SSRN.

\subsection{Completed Work}

All experimental work was finished at the time of this report.

\begin{center}
\begin{tabular}{lp{8cm}}
\toprule
Category & Status \\
\midrule
90-run experiment       & Complete: 30 runs each for A\_BASELINE, B\_CACHE, C\_FULL \\
Data collection         & Complete: all 90 rows validated across 33 columns \\
Thermal hypothesis      & Refuted: CPU constant at 27\textdegree{}C \\
Variance analysis       & Complete: OS scheduling identified as 60--70\% CV source \\
Root cause analysis     & Complete: 0\% algorithm CV vs.\ 70\% environmental CV \\
Convergence proof       & Complete: C\_FULL improvement $-25.4\%$ documented \\
Confidence bounds       & Complete: 95\% CI established per configuration \\
Paper draft             & Complete (95\%): citations and figures outstanding \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Outstanding at Time of Report}

At the November 2025 report date, the following remained:

\begin{itemize}
  \item 25--30 citations to be integrated (estimated 3--4 hours).
  \item Five publication-quality figures to be created
        (convergence curves, variance decomposition, confidence bounds,
        state-machine diagram, architecture overview).
  \item Discussion section expansion addressing practical implications
        and threats to validity.
  \item Formal appendix with extended proofs of the four theorems.
  \item Venue-specific formatting; IEEE TSE was the primary candidate.
\end{itemize}

The report assessed overall readiness at 75\%, with no critical blockers
identified.

\subsection{Data Assets}

The raw dataset consisted of \texttt{experiment\_results.csv}: 91 rows
(header plus 90 data rows), 33 columns, validated and available for
independent reproduction via the published analysis scripts
(\texttt{analyze\_thermal\_hypothesis.py},
\texttt{analyze\_variance\_sources.py}).
