%% SCRAP: archive/quality/regression/opp-02-report
%% SOURCE: docs/working/archive/quality/regression/opp-02-report.md
%% STATUS: HISTORICAL
%% FITS: none
%% EDITORIAL: lifted — prose rewritten to press voice

\section*{OPP~\#2: Rolling Window Width Optimization Report (November 2025)}

Optimization Opportunity~2 tested three rolling window width configurations
(2048, 4096, and 8192 entries) to identify the statistically optimal window
size for adaptive decay-slope inference. The experiment executed 180 runs
(60 per configuration) using the deterministic FORTH-79 test workload and the
regression detection framework established for the OPP series.

\subsection*{Results}

\begin{center}
\begin{tabular}{lccc}
\toprule
Configuration & Mean Time (ns) & vs.\ Baseline & Recommendation \\
\midrule
\texttt{WINDOW\_SIZE\_2048} & 5{,}432{,}109 & $-8.67\%$ & consider \\
\texttt{WINDOW\_SIZE\_4096} & 5{,}456{,}664 & $-8.26\%$ & not recommended \\
\texttt{WINDOW\_SIZE\_8192} & 5{,}310{,}504 & $-10.72\%$ & \textbf{lock} \\
\bottomrule
\end{tabular}
\end{center}

The OPP~\#1 baseline was 5{,}947{,}974~ns (decay slope 0.2, default window).

\subsection*{Regression Analysis}

No Type~1, 2, 4, 5, or 6 regressions were detected. All three configurations
produced Type~3 tail-latency warnings (P99 elevated 40--55\% above baseline);
analysis attributed this to increased inference computation during heartbeat
ticks at larger window sizes, not to a true regression in optimization quality.
The mean improvements (8--11\%) were judged to outweigh the P99 degradation.
\texttt{WINDOW\_SIZE\_4096} was Pareto-dominated and eliminated.

\subsection*{Lock Decision}

\texttt{ROLLING\_WINDOW\_SIZE = 8192} was locked for OPP~\#3 onward,
combining with the OPP~\#1 lock (\texttt{DECAY\_SLOPE\_Q48} $\approx 0.2$)
to form the parameter baseline for subsequent experiments. Cumulative
improvement at this point: $-10.72\%$ over the original baseline.

%% TODO(bob): Confirm whether OPP #3--5 results exist and where they are archived.
