\begingroup

\section*{Bibliography}
\addcontentsline{toc}{section}{Bibliography}

\begin{thebibliography}{99}
	
	\bibitem{knuthMMIX}
	Knuth, D. \textit{MMIXware: A RISC Computer for the Third Millennium}. Springer, 1999.
	
	\bibitem{l4Spec}
	Liedtke, J. ``On Microkernel Construction.'' \textit{SOSP}, 1995.
	
	\bibitem{seL4}
	Klein, G. et al. ``seL4: Formal Verification of an OS Kernel.'' \textit{SOSP}, 2009.
	
	\bibitem{fiascoOC}
	Fiasco.OC Microkernel Architecture Reference, TU Dresden.
	
	\bibitem{forthInventors}
	Moore, C. \& Pountain, G. ``The Evolution of the Forth Programming Language.'' \textit{Byte}, 1980.
	
	\bibitem{smalltalkBlueBook}
	Goldberg, A. \& Robson, D. \textit{Smalltalk-80: The Language and its Implementation}. Addison-Wesley, 1983.
	
	\bibitem{jvmSpec}
	Lindholm, T. \& Yellin, F. \textit{Java Virtual Machine Specification}. Addison-Wesley, 1999.
	
	\bibitem{feedbackSystems}
	Åström, K. \& Murray, R. \textit{Feedback Systems}. Princeton University Press, 2008.
	
	\bibitem{montgomeryDoE}
	Montgomery, D. \textit{Design and Analysis of Experiments}. Wiley, 2019.
	
	\bibitem{boxHunter}
	Box, G., Hunter, W., Hunter, J. \textit{Statistics for Experimenters}. Wiley, 2005.
	
	\bibitem{bda}
	Gelman, A. et al. \textit{Bayesian Data Analysis}. Chapman \& Hall, 2013.
	
	\bibitem{startingForth}
	Brodie, L. \textit{Starting Forth}. Prentice-Hall, 1981.
	
	\bibitem{thinkingForth}
	Brodie, L. \textit{Thinking Forth}. Prentice-Hall, 1984.
	
	\bibitem{cacheOptimizations}
	Hennessy, J. \& Patterson, D. \textit{Computer Architecture: A Quantitative Approach}. Morgan Kaufmann, 2018.
	
	\bibitem{signalProcessing}
	Oppenheim, A. \& Schafer, R. \textit{Discrete-Time Signal Processing}. Prentice Hall, 2009.
	
\end{thebibliography}

\endgroup
