%% SCRAP: archive/quality/validation/comprehensive-physics-validation
%% SOURCE: docs/working/archive/quality/validation/comprehensive-physics-validation.md
%% STATUS: HISTORICAL
%% FITS: none
%% EDITORIAL: lifted — prose rewritten to press voice

\section*{Comprehensive Physics Engine Validation Protocol (November 2025)}

This protocol defined the rigorous experimental design for validating the
StarForth physics engine across three cumulative feature configurations.
The experiment was executed as specified and its results published in the
\textsc{ssrn} paper. The protocol is retained as a provenance record documenting
the experimental methodology.

\subsection*{Experimental Design}

Three binary configurations were tested, 30 runs each (90 total):

\begin{itemize}
  \item \textbf{A\_BASELINE}: \texttt{ENABLE\_HOTWORDS\_CACHE=0},
        \texttt{ENABLE\_PIPELINING=0}. Control condition.
  \item \textbf{B\_CACHE}: \texttt{ENABLE\_HOTWORDS\_CACHE=1},
        \texttt{ENABLE\_PIPELINING=0}. Isolates cache impact.
  \item \textbf{C\_FULL}: \texttt{ENABLE\_HOTWORDS\_CACHE=1},
        \texttt{ENABLE\_PIPELINING=1}. Combined physics optimization.
\end{itemize}

Each run executed a three-phase benchmark: warmup (100{,}000 iterations,
discarded), statistics reset, then measurement (100{,}000 iterations, recorded).
This design eliminates cold-start artifacts and measures steady-state performance.

Run order was randomized across all 90 trials using a shuffled test matrix,
distributing configuration-switch costs evenly and eliminating temporal bias.
Only three binary rebuilds were required despite the interleaved execution
sequence.

\subsection*{Statistical Analysis}

Per-configuration analysis computed mean, median, standard deviation, and
Bayesian 95\% and 99\% credible intervals. Comparative analysis isolated the
incremental contribution of each optimization layer. The coefficient of variation
(CV) served as the reproducibility metric; a CV below 5\% was defined as
indicative of a highly deterministic physics model.

\subsection*{Outcome}

The experiment confirmed the null hypothesis was false for both comparisons.
Cache hit rate exceeded the 20\% minimum threshold; speedup under B\_CACHE
was statistically significant; algorithmic CV under B\_CACHE was 0\%;
OS-scheduling variance accounted for 60--70\% of total variance. Full results
appear in the published \textsc{ssrn} paper.
