%% SCRAP: archive/quality/validation/phase-1-protocol
%% SOURCE: docs/working/archive/quality/validation/phase-1-protocol.md
%% STATUS: HISTORICAL
%% FITS: none
%% EDITORIAL: lifted — prose rewritten to press voice

\section{Phase 1 Governance Validation Protocol (November 2025)}

This protocol, prepared 3 November 2025, defined the acceptance test
for the Phase 1 GitHub governance workflow system. It verified that all
document types routed correctly, approval chains triggered, and backlog
access control was enforced. The system being tested has since been
superseded; the protocol is preserved as a historical audit record.

\subsection{Scope}

The protocol covered ten validation domains: template recognition across
16 document types, submission workflows, approval chains, PM routing
decisions, backlog entry, vault routing to \texttt{in\_basket}, gatekeeper
enforcement against unauthorized access, security event logging to
\texttt{SEC\_LOG.adoc}, label management, and error handling.

\subsection{Test Structure}

Seven test suites organized validation:

\begin{description}
  \item[Suite 1 — Template Recognition.]
        All 16 document types (ECO, CAPA, CER, DHR, DWG, ENG, SEC, IR,
        VAL, DTA, DMR, ART, MIN, REL, RMP, and others) were submitted via
        GitHub issue templates and checked for correct label application.
  \item[Suite 2 — Approval Workflows.]
        CAPA QA approval, PM routing decisions, and CER three-phase
        sequential approval (Protocol, Results, Report) were verified end
        to end.
  \item[Suite 3 — Routing and Backlog Entry.]
        Approved documents were routed to the development backlog or
        governance vault. Reference documents (DHR, DMR, etc.) auto-vaulted
        without manual approval.
  \item[Suite 4 — Gatekeeper Enforcement.]
        Unauthorized backlog access attempts were blocked; authorized
        paths and owner (rajames440) direct access were permitted.
  \item[Suite 5 — Security Audit Logging.]
        All bypass attempts were verified in \texttt{SEC\_LOG.adoc} with
        timestamps, actor usernames, and severity labels. Incident Report
        (IR) issues were auto-created for each violation.
  \item[Suite 6 — Label Management.]
        Type label exclusivity and status label progression were verified.
  \item[Suite 7 — Error Handling.]
        Incomplete template detection and workflow feedback comment quality
        were assessed.
\end{description}

\subsection{Pass Criteria}

The system passed if all 16 document types were recognizable, all approval
workflows triggered correctly, documents routed to the correct destination,
the gatekeeper blocked unauthorized access while permitting authorized
access, all events logged to \texttt{SEC\_LOG.adoc}, and user feedback was
clear and actionable.

\subsection{Execution Plan}

The protocol was designed for four-day execution (7--11 hours total):
Day 1 for templates and approvals, Day 2 for routing and backlog, Day 3
for security and enforcement, Day 4 for labels and feedback.
