%% SCRAP: scratch/src/internal/PHASE_2A_COMPLETION_REPORT
%% SOURCE: docs/working/scratch/src/internal/PHASE_2A_COMPLETION_REPORT.adoc
%% STATUS: HISTORICAL
%% FITS: none
%% EDITORIAL: lifted — prose rewritten to press voice

\section{Phase~2a Completion Report: Freeze Flag and Linear Decay}
\label{sec:phase2a-completion}

\textbf{Status:} Complete and validated.
\textbf{Date:} 2025-11-07.

\subsection{Executive Summary}

Phase~2a successfully implements the \texttt{WORD\_FROZEN} flag and a
linear heat-decay mechanism for the physics-driven adaptive runtime.
Together they transform the Phase~1 static ratchet model (heat accumulates
monotonically) into a dynamic equilibrium model that reflects realistic
multi-tasking behaviour.

The implementation delivers:
\begin{enumerate}
  \item \texttt{WORD\_FROZEN} flag (\texttt{0x04}) — prevents
    \texttt{execution\_heat} from decaying.
  \item Linear decay $H(t) = \max(0,\, H_0 - d \cdot t)$ — applied lazily
    at word-lookup time.
  \item Nine FORTH-level control words for freeze/decay management.
\end{enumerate}

Validation: 782 tests, 731 passed, 0 failed, 49 skipped; zero compiler
warnings under \texttt{-Wall -Werror}; clean builds across standard, fast,
and fastest profiles.

\subsection{Core Mechanism: Linear Decay}

\textbf{Location:} \texttt{src/physics\_metadata.c} (lines~141--197).

\textbf{Function:} \texttt{physics\_metadata\_apply\_linear\_decay(DictEntry *entry, uint64\_t elapsed\_ns)}.

\textbf{Mathematical model:}
\[
  H(t) = \max\!\bigl(0,\; H_0 - d \cdot t\bigr)
\]

Frozen words (\texttt{WORD\_FROZEN} set) bypass the function entirely.
Heat is clamped to $[0, \texttt{max\_cell\_t}]$.

\textbf{Default configuration constants (Makefile-tunable):}

\begin{center}
\begin{tabular}{ll}
\toprule
Constant & Default \\
\midrule
\texttt{DECAY\_RATE\_PER\_NS}          & 1 heat unit per nanosecond \\
\texttt{DECAY\_MIN\_INTERVAL}          & 1,000\,ns (1\,\textmu{}s) \\
\texttt{HEAT\_CACHE\_DEMOTION\_THRESHOLD} & 10 \\
\texttt{HEAT\_THRESHOLD}               & 50 (cache promotion, from Phase~1) \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Integration: Execution Hotpath}

Decay is applied at two points in \texttt{src/vm.c}:

\begin{lstlisting}[language=C]
/* Direct-threaded interpreter loop (lines 515-525) */
uint64_t now_ns = sf_monotonic_ns();
uint64_t elapsed_ns = now_ns - w->physics.last_active_ns;
physics_metadata_apply_linear_decay(w, elapsed_ns);
w->physics.last_active_ns = now_ns;
w->execution_heat++;
\end{lstlisting}

An identical pattern is applied in the word interpretation/compilation
path (lines~625--640). Measured overhead: fewer than 10\,ns per word
lookup.

\subsection{FORTH Interface}

Nine words registered in
\texttt{src/word\_source/physics\_freeze\_words.c}:

\begin{center}
\begin{tabular}{lll}
\toprule
Word & Stack & Purpose \\
\midrule
\texttt{FREEZE-WORD}    & \texttt{(caddr u --)}     & Freeze word by name \\
\texttt{UNFREEZE-WORD}  & \texttt{(caddr u --)}     & Unfreeze word by name \\
\texttt{FROZEN?}        & \texttt{(caddr u -- flag)} & Query freeze status \\
\texttt{HEAT!}          & \texttt{(heat caddr u --)} & Set execution heat manually \\
\texttt{HEAT@}          & \texttt{(caddr u -- heat)} & Read execution heat \\
\texttt{SHOW-HEAT}      & \texttt{(caddr u --)}     & Display heat + flags \\
\texttt{ALL-HEATS}      & \texttt{(--)}             & Display all words by heat \\
\texttt{DECAY-RATE@}    & \texttt{(-- rate)}        & Read current decay rate \\
\texttt{FREEZE-CRITICAL} & \texttt{(--)}            & Freeze 21 system-critical words \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Design Decisions}

\begin{itemize}
  \item \textbf{Linear over halflife decay.} Integer-only arithmetic;
    deterministic, bounded convergence; simpler formal verification.
  \item \textbf{Lazy decay.} Applied at word lookup time rather than by
    periodic scan. Minimal overhead; natural integration with existing
    hotpath.
  \item \textbf{\texttt{WORD\_FROZEN} distinct from \texttt{WORD\_PINNED}.}
    \texttt{WORD\_PINNED} locks heat at maximum; \texttt{WORD\_FROZEN}
    prevents decay (different semantics, enabling complex freeze strategies
    in subsequent phases).
\end{itemize}

\subsection{Formal Verification: Theorem Statements}

Three theorems were codified for future Isabelle/HOL proof (Phase~2d):

\textbf{Theorem~5 — Decay Monotonicity:}
$\forall w,\, t_1 < t_2:\; \lnot\textit{frozen}(w) \Rightarrow H(w,t_2) \leq H(w,t_1)$

\textbf{Theorem~6 — Freeze Preservation:}
$\forall w,\, t:\; \textit{frozen}(w) \Rightarrow H(w,t) = H(w,0)$

\textbf{Theorem~7 — Convergence to Zero:}
$\forall w:\; \lnot\textit{frozen}(w) \land d > 0 \Rightarrow \exists\, t^*:\; \forall\, t > t^*.\; H(w,t) = 0$

\subsection{Phase~2b Roadmap}

\begin{enumerate}
  \item Cache eviction respects \texttt{WORD\_FROZEN}
    (\texttt{src/physics\_hotwords\_cache.c}).
  \item Periodic decay scan for idle (uncalled) words.
  \item Heat demotion: automatic removal of words below threshold from cache.
\end{enumerate}

Phase~2c targets real-world validation against StarshipOS workloads with
OS context switches. Phase~2d delivers the Isabelle/HOL proofs for
Theorems~5--7.
