%% SCRAP: scratch/src/internal/PHASE_2_IMPLEMENTATION_SUMMARY
%% SOURCE: docs/working/scratch/src/internal/PHASE_2_IMPLEMENTATION_SUMMARY.adoc
%% STATUS: HISTORICAL
%% FITS: none
%% EDITORIAL: lifted — prose rewritten to press voice

\section{Phase~2 Implementation Summary and Action Items}
\label{sec:phase2-impl-summary}

\textbf{Status:} Specification complete; ready for code implementation.
\textbf{Date:} 2025-11-07.

This document summarises the deliverables produced at the close of the
Phase~2 specification session and lists the four implementation tasks
required for Phase~2a.

\subsection{Deliverables Completed}

The Phase~2 specification document (\texttt{PHASE\_2\_PHYSICS\_DECAY\_AND\_FREEZE.adoc})
was completed, covering: problem statement, two technical designs (freeze flag
and linear decay mechanism), mathematical comparison of halflife versus linear
models, implementation recommendation, integration points, FORTH API, testing
strategy, four-week deployment roadmap, and Isabelle/HOL targets for
Theorems~5--7.

\subsection{Implementation Checklist: Phase~2a}

\textbf{Task~1 — Add \texttt{WORD\_FROZEN} flag to \texttt{include/vm.h}.}
Estimated effort: 15 minutes.

\begin{lstlisting}[language=C]
#define WORD_FROZEN   0x04  /* heat does not decay */
\end{lstlisting}

\textbf{Task~2 — Implement linear decay function in
\texttt{src/physics\_metadata.c}.}
Estimated effort: 30 minutes.

\begin{lstlisting}[language=C]
void physics_metadata_apply_linear_decay(DictEntry *entry,
                                         uint64_t elapsed_ns) {
    if (!entry || (entry->flags & WORD_FROZEN)) return;
    if (elapsed_ns < DECAY_MIN_INTERVAL) return;

    cell_t decay = (cell_t)(elapsed_ns / NS_PER_UNIT * DECAY_RATE_PER_NS);
    if (decay >= entry->execution_heat)
        entry->execution_heat = 0;
    else
        entry->execution_heat -= decay;
}
\end{lstlisting}

\textbf{Task~3 — Integrate decay into word-lookup hotpath in
\texttt{src/vm.c}.}
Estimated effort: 45 minutes. Insert after dictionary lookup and before
heat accumulation.

\textbf{Task~4 — Create FORTH interface words.}
Estimated effort: 30 minutes. Words required: \texttt{FREEZE-WORD},
\texttt{UNFREEZE-WORD}, \texttt{FROZEN?}.

\textbf{Total estimated effort for Phase~2a:} 3--4 hours (including testing).

\subsection{Files to Modify}

\begin{center}
\begin{tabular}{lll}
\toprule
File & Change & Purpose \\
\midrule
\texttt{include/vm.h}               & Add constant   & \texttt{WORD\_FROZEN} flag \\
\texttt{src/physics\_metadata.c}    & Add function   & Implement decay \\
\texttt{src/physics\_metadata.h}    & Add declaration & Declare decay function \\
\texttt{src/vm.c}                   & Modify hotpath & Call decay before heat++ \\
\texttt{src/word\_source/physics\_freeze\_words.c} & New file & Freeze control words \\
\texttt{include/word\_registry.h}   & Add entries    & Register new words \\
\texttt{src/test\_runner/modules/test\_physics\_decay.c} & New file & Unit tests \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Key Design Decisions}

\begin{center}
\begin{tabular}{lll}
\toprule
Decision & Rationale & Status \\
\midrule
Linear decay (not halflife) & Integer arithmetic, deterministic & Chosen \\
Lazy decay (on access)      & Minimal overhead, natural integration & Chosen \\
\texttt{WORD\_FROZEN} distinct from \texttt{WORD\_PINNED} & Different semantics & Chosen \\
Configuration via \texttt{\#define} + Makefile & Standard pattern & Chosen \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Success Criteria for Phase~2a}

Must have: \texttt{WORD\_FROZEN} compiles cleanly; decay function passes
unit tests; decay integrated in hotpath; \texttt{FREEZE-WORD} and
\texttt{UNFREEZE-WORD} work interactively; no regression (all prior
tests pass).

Nice to have: performance regression under 10\%; integration test
demonstrating frozen words retain heat; documentation complete.

Explicitly excluded from Phase~2a: cache eviction logic changes;
periodic decay scan; Isabelle/HOL proofs; StarshipOS real-world validation.
