%% SCRAP: scratch/src/internal/PHASE_2_PHYSICS_DECAY_AND_FREEZE
%% SOURCE: docs/working/scratch/src/internal/PHASE_2_PHYSICS_DECAY_AND_FREEZE.adoc
%% STATUS: HISTORICAL
%% FITS: dev-guide/ch-physics if still accurate; assess honestly
%% EDITORIAL: lifted — prose rewritten to press voice. Design options
%%   section preserved; implementation is now the canonical behaviour
%%   described in Phase 2a completion report.

\section{Phase~2: Physics Model Extensions — Decay and Freeze}
\label{sec:phase2-decay-freeze}

\textbf{Status at time of writing:} Design phase, ready for implementation.
\textbf{Date:} 2025-11-07.

This document specifies the freeze flag and decay mechanism that extend the
Phase~1 physics model. Phase~2a has since been completed; this document is
retained as the design rationale record.

\subsection{Problem Statement}

Phase~1 tracks \texttt{execution\_heat} as a monotonically increasing
counter. When the StarForth VM is preempted and a new task begins execution,
the old heat values remain in the dictionary and misrepresent the new task's
access patterns. Words that were hot for the previous task pollute the
hot-words cache, causing cache misses for the new task's actual hot path.

A decay mechanism addresses this by dissipating heat over time, and a freeze
mechanism ensures that genuinely cross-context critical words
(e.g.\ \texttt{DUP}, \texttt{DROP}, \texttt{SWAP}) remain cached regardless.

\subsection{Word Flag Extension}

A new flag is added to the \texttt{DictEntry} flag byte:

\begin{lstlisting}[language=C]
#define WORD_FROZEN   0x04  /* execution_heat does not decay */
\end{lstlisting}

The complete flag layout is:

\begin{center}
\begin{tabular}{ll}
\toprule
Bit & Flag \\
\midrule
\texttt{0x80} & \texttt{WORD\_IMMEDIATE} \\
\texttt{0x40} & \texttt{WORD\_HIDDEN} \\
\texttt{0x20} & \texttt{WORD\_SMUDGED} \\
\texttt{0x10} & \texttt{WORD\_COMPILED} \\
\texttt{0x08} & \texttt{WORD\_PINNED} \\
\texttt{0x04} & \texttt{WORD\_FROZEN} \\
\texttt{0x02} & (reserved) \\
\texttt{0x01} & (reserved) \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Decay Models Considered}

\subsubsection{Halflife (Exponential) Decay}

\[
  H(t) = H_0 \cdot 2^{-t/\tau}
\]

Smooth, asymptotic approach to zero. Requires floating-point arithmetic
(\texttt{pow()}). Heat never reaches zero exactly. Considered and rejected
for Phase~2a.

\subsubsection{Linear Decay (Selected)}

\[
  H(t) = \max\!\bigl(0,\; H_0 - d \cdot t\bigr)
\]

Pure integer subtraction. Heat reaches zero in finite time $t^* = H_0/d$.
Predictable and bounded. Selected for Phase~2a.

\begin{center}
\begin{tabular}{lcc}
\toprule
Criterion & Halflife & Linear \\
\midrule
Integer arithmetic       & No  & Yes \\
Bounded convergence      & No  & Yes \\
Simplicity               & No  & Yes \\
Formal verifiability     & Harder & Easier \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Integration Points}

Decay is applied at word-lookup time before heat accumulation (lazy strategy).
A check prevents application for elapsed times below \texttt{DECAY\_MIN\_INTERVAL}
(default 1\,\textmu{}s) to avoid noise from rapid successive calls.

Cache eviction logic (Phase~2b) must also check the \texttt{WORD\_FROZEN}
flag and skip frozen candidates:

\begin{lstlisting}[language=C]
if (candidate->flags & WORD_FROZEN) continue;
\end{lstlisting}

\subsection{Configuration}

All decay parameters are tunable via Makefile override:

\begin{lstlisting}[language=bash]
# Aggressive decay
make DECAY_RATE_PER_NS=2

# Conservative decay
make DECAY_RATE_PER_NS=0.5 HEAT_THRESHOLD=25
\end{lstlisting}

\subsection{Formal Verification Targets}

Three theorems for Isabelle/HOL proof (Phase~2d):

\begin{description}
  \item[Theorem~5 — Decay Monotonicity]
    For non-frozen words, heat is non-increasing over time.
  \item[Theorem~6 — Freeze Preservation]
    Frozen words maintain constant heat for all time.
  \item[Theorem~7 — Convergence]
    Non-frozen words with positive decay rate reach zero in finite time.
\end{description}

\subsection{Deployment Roadmap}

\begin{description}
  \item[Phase~2a] Freeze flag + linear decay (complete as of 2025-11-07).
  \item[Phase~2b] Cache eviction respects frozen flag; heat demotion.
  \item[Phase~2c] StarshipOS real-world validation with OS context switches.
  \item[Phase~2d] Isabelle/HOL proofs for Theorems~5--7.
\end{description}
