%% SCRAP: scratch/src/internal/PHYSICS_IMPLEMENTATION_STATUS
%% SOURCE: docs/working/scratch/src/internal/PHYSICS_IMPLEMENTATION_STATUS.adoc
%% STATUS: HISTORICAL
%% FITS: dev-guide/ch-physics
%% EDITORIAL: lifted — prose rewritten to press voice

\section{Physics Engine Implementation Status: POSIX vs.\ L4Re Porting Strategy}

This document records the implementation status of the physics engine at
the time the porting strategy was being finalized. It is HISTORICAL;
the current state of the codebase should be verified directly.

\subsection{What Is Wired}

Both word-execution paths call physics metadata updates:

\begin{itemize}
  \item \textbf{Outer interpreter} (\texttt{src/vm.c:533}) — direct word
        execution: calls \texttt{physics\_metadata\_touch()} after each
        primitive dispatch.
  \item \textbf{Colon word executor} (\texttt{src/vm.c:456}) — nested word
        calls: same call after each colon-definition word dispatch.
\end{itemize}

Both paths update \texttt{temperature\_q8} (exponential moving average from
entropy) and \texttt{last\_active\_ns} (monotonic timestamp).

Dictionary initialization calls \texttt{physics\_metadata\_init()} and
\texttt{physics\_metadata\_apply\_seed()}, pre-seeding 20 high-impact words
(IF, LOOP, EMIT, BLOCK, SAVE-BUFFERS, etc.) with domain-knowledge priors.

\subsection{What Is Incomplete}

\begin{description}
  \item[POSIX/L4Re abstraction pattern.]
        The physics runtime uses inline \texttt{\#ifdef \_\_l4\_\_}
        conditionals rather than the vtable pattern established by
        \texttt{include/platform\_time.h}. Recommended refactoring: extract
        POSIX code to \texttt{src/platform/linux/snapshot.c} and the L4Re
        stub to \texttt{src/platform/l4re/snapshot.c}, with a vtable
        dispatcher in \texttt{src/physics\_runtime.c}.

  \item[Periodic host snapshot capture.]
        \texttt{physics\_host\_snapshot()} is defined but never called
        automatically. No heartbeat mechanism invokes it; the analytics heap
        is allocated but unused in normal operation.

  \item[Temperature decay model.]
        \texttt{temperature\_q8} is updated on execution but never decayed
        for inactive words. The exponential moving average with configurable
        half-life remains unimplemented.
\end{description}

\subsection{POSIX Implementation Status}

The POSIX path is production-ready and robust. Captured signals include CPU
count (\texttt{sysconf}), scheduler policy and priority (\texttt{sched\_*}),
time quantum (\texttt{sched\_rr\_get\_interval}), load average
(\texttt{getloadavg}), PSI metrics (\texttt{/proc/pressure/}), \texttt{/proc/stat}
jiffies, and cgroup v2 (\texttt{cpu.stat}, \texttt{memory.current}).
Missing interfaces result in \texttt{flags = 0} for the affected category;
no crashes.

\subsection{L4Re Implementation Status}

The L4Re path is a stub: monotonic and real-time timestamps via the platform
time abstraction work correctly; CPU count is hardcoded to 1 (KIP query
deferred). Scheduler info via \texttt{l4\_scheduler\_info()} and IO server
capabilities are planned for later phases.

\subsection{Recommended Next Steps}

\begin{enumerate}
  \item Verify temperature updates in \texttt{starforth\_words\_test.c}:
        define a word, execute it $N$ times, inspect \texttt{temperature\_q8}.
  \item Add periodic host snapshot capture to the REPL loop or VM tick.
  \item Implement temperature decay (exponential moving average with
        configurable half-life) in \texttt{src/physics\_metadata.c}.
  \item Refactor to the vtable pattern once the logic is proven.
\end{enumerate}
