%% SCRAP: scratch/src/internal/PHYSICS_HOTWORDS_CACHE_INTEGRATION
%% SOURCE: docs/working/scratch/src/internal/PHYSICS_HOTWORDS_CACHE_INTEGRATION.adoc
%% STATUS: CURRENT
%% FITS: dev-guide/ch-physics
%% EDITORIAL: lifted — prose rewritten to press voice

\section{Hot-Words Cache Integration Guide}

The hot-words cache (\texttt{ENABLE\_HOTWORDS\_CACHE=1}, default on) is a
frequency-driven dictionary acceleration feature. It uses per-word execution
heat to maintain a small LRU cache of the most frequently executed words and
reorders dictionary buckets by execution frequency to reduce average search
depth.

\textbf{Expected impact:} 2--4$\times$ speedup on hot-word lookups (IF, THEN,
DO, LOOP, etc.) under typical FORTH workloads.

\subsection{Architecture}

\subsubsection{Cache Structure}

The \texttt{HotwordsCache} struct (32 entries by default) is embedded in the
\texttt{VM} struct. The lookup path:

\begin{enumerate}
  \item Check \texttt{cache[0..cache\_count]} — cache hit returns immediately.
  \item Search the bucket for the name — if found and hot, promote to cache.
  \item LRU eviction if the cache is full.
\end{enumerate}

Key parameters (in \texttt{include/physics\_hotwords\_cache.h}):

\begin{center}
\begin{tabular}{lll}
\toprule
Constant & Default & Effect \\
\midrule
\texttt{HOTWORDS\_CACHE\_SIZE}                 & 32  & Cache slots; larger = more hits, slower scan \\
\texttt{HOTWORDS\_ENTROPY\_THRESHOLD}          & 50  & Executions required for cache admission \\
\texttt{HOTWORDS\_ENTROPY\_DELTA\_THRESHOLD}   & 100 & Delta required to trigger bucket reordering \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Benchmark Words}

After integration, the following FORTH words are available for measurement:

\begin{center}
\begin{tabular}{lp{8cm}}
\toprule
Word & Purpose \\
\midrule
\texttt{1000 BENCH-DICT-LOOKUP} & Run 1{,}000 dictionary lookups and show timing \\
\texttt{PHYSICS-CACHE-STATS}    & Display hit rate, promotion count, eviction count \\
\texttt{PHYSICS-TOGGLE-CACHE}   & Enable/disable cache at runtime for A/B testing \\
\texttt{PHYSICS-RESET-STATS}    & Reset all statistics for a clean measurement \\
\texttt{PHYSICS-BUILD-INFO}     & Show current build configuration \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Before/After Measurement Protocol}

\begin{lstlisting}[language=bash]
# Baseline (cache disabled)
make clean && make ENABLE_HOTWORDS_CACHE=0 test

# With cache enabled (default)
make clean && make ENABLE_HOTWORDS_CACHE=1 test
\end{lstlisting}

In the REPL:

\begin{lstlisting}[language=forth]
PHYSICS-RESET-STATS
100000 BENCH-DICT-LOOKUP
PHYSICS-CACHE-STATS
\end{lstlisting}

\subsection{Expected Results}

\begin{center}
\begin{tabular}{lll}
\toprule
Scenario & Cache Hit Rate & Speedup \\
\midrule
Hot-word heavy (IF, THEN, DO, LOOP) & 40--60\% & 2--4$\times$ \\
Balanced workload                   & 20--30\% & 1.3--1.8$\times$ \\
Cold-word heavy                     & 5--15\%  & 1.1--1.2$\times$ \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Memory overhead:} approximately 512 bytes (pointer array plus stats
struct). \textbf{CPU overhead:} negligible — array scan is substantially
faster than bucket search.

\subsection{References}

\begin{itemize}
  \item \texttt{src/physics\_hotwords\_cache.c} —
        cache implementation
  \item \texttt{include/physics\_hotwords\_cache.h} —
        constants and struct definitions
  \item \texttt{src/word\_source/physics\_benchmark\_words.c} —
        benchmark word definitions
  \item \texttt{src/physics\_metadata.c} —
        metrics collection
\end{itemize}
