%% SCRAP: archive/DOCUMENTATION_AUDIT
%% SOURCE: docs/working/archive/DOCUMENTATION_AUDIT.md
%% STATUS: HISTORICAL
%% FITS: none
%% EDITORIAL: lifted — prose rewritten to press voice

\section{Documentation Audit: December 2025}

A full audit of the StarForth documentation directory conducted on 2025-12-14
produced the following findings. The record is preserved here as a historical
milestone; subsequent reorganisation phases have addressed the majority of
issues identified.

\subsection{Executive Summary}

At the time of the audit, the documentation was well-organised into a
numbered directory structure (\texttt{01}--\texttt{07}) with strong coverage
of experiments, architecture, and quality processes. The newly added Hardware
Abstraction Layer (HAL) documentation set the quality bar for the project.
Six key issues were identified:

\begin{itemize}
  \item Missing \texttt{README.md} files in six of seven numbered directories.
  \item Inconsistent metadata (no author, date, or status fields on older documents).
  \item Top-level clutter: thirteen or more loose files in \texttt{docs/}.
  \item Unclear articulation of the StarKernel and StarshipOS vision in existing documents.
  \item The project \texttt{CLAUDE.md} did not yet reference the HAL architecture.
  \item No single architecture overview tying all subsystems together.
\end{itemize}

\subsection{Prioritised Recommendations}

The audit assigned priority as follows.

\textbf{High priority:}
\begin{itemize}
  \item Add missing \texttt{README} files to all numbered directories.
  \item Update \texttt{CLAUDE.md} to reflect HAL and StarKernel vision.
  \item Create \texttt{docs/03-architecture/README.md} as a system overview.
\end{itemize}

\textbf{Medium priority:}
\begin{itemize}
  \item Create \texttt{docs/03-architecture/OVERVIEW.md}.
  \item Consolidate and relocate top-level loose files.
  \item Add \texttt{CONTRIBUTING.md}.
  \item Add metadata headers to existing documents.
\end{itemize}

\textbf{Low priority:}
\begin{itemize}
  \item Create a project glossary.
  \item Add architecture diagrams (PlantUML or Mermaid, exported to SVG).
\end{itemize}

\subsection{HAL Documentation Assessment}

The HAL documentation set added in late 2025 was rated excellent: six
documents totalling approximately twenty thousand words, covering overview,
interface contracts, platform implementation, incremental migration strategy,
and StarKernel-specific integration details. Its \texttt{README.md} navigation
hub was recommended as the template for all other subsection indices.

\subsection{Success Metrics}

Documentation improvement is complete when:
\begin{enumerate}
  \item Every numbered directory contains a \texttt{README}.
  \item The StarForth $\to$ StarKernel $\to$ StarshipOS vision is unambiguous in top-level documents.
  \item New contributors can navigate the documentation without assistance.
  \item All major subsystems have an overview document and detailed subdocuments.
  \item The top-level \texttt{docs/} directory contains fewer than five loose files.
  \item All cross-references resolve correctly.
  \item Consistent style and metadata headers appear across all documents.
\end{enumerate}
