%% SCRAP: scratch/ci-cd/GITHUB_ISSUES_GOVERNANCE_FLOW
%% SOURCE: docs/working/scratch/ci-cd/GITHUB_ISSUES_GOVERNANCE_FLOW.adoc
%% STATUS: OBSOLETE
%% FITS: none
%% EDITORIAL: lifted — prose rewritten to press voice

\section{GitHub Issues and Governance Integration}

This note records the architectural decision governing the relationship between GitHub
issue types (CAPA, ECR, ECO) and the StarForth formal governance system.

\subsection{Architecture Decision}

GitHub Issues follow a development workflow and do not feed the \texttt{in\_basket}
governance intake pathway directly. The project uses a two-path model.

\paragraph{Development governance (GitHub Issues).}
Users and developers submit issues using typed templates. Automation validates and
labels each submission. A Kanban board tracks progress through the developer workflow.
Resolution is a merged pull request.

\paragraph{Operational governance (CI/CD artifacts).}
CI/CD pipelines generate the artifacts that constitute governance evidence: test
results, benchmark data, build manifests, and SBOM outputs. These artifacts flow
automatically into \texttt{in\_basket/} for QA triage and eventual disposition to
the Vault.

The two paths are connected by links rather than duplication. A Vault record for a
resolved CAPA includes the GitHub issue URL, the PR number, and a reference to the
corresponding \texttt{in\_basket} test result, providing complete traceability without
redundant data entry.

\subsection{Exceptions}

Direct export of GitHub Issues to \texttt{in\_basket} is reserved for specific
circumstances: external compliance certification requirements, off-platform archival
during a migration away from GitHub, or an auditor request for a consolidated
point-in-time record.

%% TODO(bob): assess whether any certification body has since required direct issue export.
