%% SCRAP: scratch/ci-cd/GOVERNANCE_INPUT_AUDIT
%% SOURCE: docs/working/scratch/ci-cd/GOVERNANCE_INPUT_AUDIT.adoc
%% STATUS: HISTORICAL
%% FITS: none
%% EDITORIAL: lifted — prose rewritten to press voice

\section{Governance Input Audit (Phase 1)}

This audit assessed all controlled governance inputs against the requirement that every
document bound for the Vault must pass through \texttt{in\_basket/} as the sole gating
point. The audit was conducted on 2025-11-03 and returned a \textbf{COMPLIANT} verdict
with implementation verified for five of eight input sources and documentation complete
for the remaining three.

\subsection{Compliance Matrix}

\begin{center}
\begin{tabular}{llll}
\toprule
Input source & Artifact type & Route & Status \\
\midrule
\texttt{test-stage.yml} & Test results & \texttt{in\_basket/Test\_Results/} & Implemented \\
\texttt{qual-stage.yml} & Benchmarks & \texttt{in\_basket/Performance\_Data/} & Implemented \\
\texttt{qual-stage.yml} & SBOM & \texttt{in\_basket/} & Implemented \\
\texttt{prod-stage.yml} & Build manifest & \texttt{in\_basket/Build\_Manifests/} & Implemented \\
Jenkins devl job & Build logs & \texttt{in\_basket/Test\_Results/} & Pending \\
Jenkins test job & Test results & \texttt{in\_basket/Test\_Results/} & Pending \\
Jenkins qual job & Benchmarks, SBOM, verification & \texttt{in\_basket/Performance\_Data/} & Pending \\
GitHub Issues & CAPA / ECR / ECO & Kanban (development) & Correct by design \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Backdoor Analysis}

The audit identified four theoretical bypass vectors and the controls in place for each.

\begin{itemize}
  \item \textbf{Direct Vault commit.} Mitigated by pull-request peer review and branch
        protection on the governance repository.
  \item \textbf{Workflow dynamic path injection.} Mitigated by code review of all
        \texttt{.github/workflows/*.yml} files; paths are hardcoded, not constructed at
        runtime.
  \item \textbf{Jenkins SSH direct write.} Mitigated by scoping SSH credentials to the
        governance repository and documenting required routing in
        \texttt{JENKINS\_GOVERNANCE\_INTEGRATION.adoc}. Risk rated Medium.
  \item \textbf{Manual operator bypass.} Mitigated by the \texttt{INTAKE\_LOG} requirement
        and the gating rules in \texttt{INTAKE\_PROCEDURES.adoc}. Risk rated Medium.
\end{itemize}

\subsection{Pending Actions}

At audit close, three Jenkins jobs required Jenkinsfile updates to route outputs through
\texttt{in\_basket/} per the documented specification. The end-to-end flow for GitHub
Actions pipelines was verified working.

%% TODO(bob): verify whether Jenkins jobs were ever updated post-audit.
