%% Vol I, Chapter 4 — Formal Verification
%% SOURCE: proof/ directory (24 .thy files — confirmed by ls proof/)
%%         .claude/CLAUDE.md (Formal Verification section)
%%         proof/StarForth_Base.thy, proof/StarForth_Correctness.thy,
%%         proof/StarForth_Concurrent.thy, proof/StarForth_Mutex.thy,
%%         proof/StarForth_Transition.thy, proof/ACL_*.thy (5 files)
%%         proof/StarForth_Loop{1..7}_*.thy (7 files)
%%         proof/StarForth_{Arithmetic,Stack,Logical,Memory,Return_Stack,Q48_16}_Words.thy

\chapter{Formal Verification}
\label{vol1:chap:proofs}

StarForth's physics-driven adaptive runtime and FORTH-79 word set are formally
verified using Isabelle/HOL. Unlike assertions or unit tests, Isabelle/HOL
produces machine-checked proofs: every theorem is mechanically verified from
first principles within a small, well-audited logical kernel, providing a level
of assurance that runtime testing alone cannot reach. The proof corpus consists
of 24 theory files covering all seven physics feedback loops, six FORTH-79 word
categories, the Q48.16 fixed-point arithmetic subsystem, concurrency and mutual
exclusion properties, and the word-level ACL security system.

\section{Proof Infrastructure}
\label{vol1:sec:proofs:infra}

All theory files reside in the \texttt{proof/} directory at the repository
root. The session is declared in \texttt{proof/ROOT}, which lists the import
order and the session name for the Isabelle build system. All 24 theory files
import either \texttt{StarForth\_Base} or a theory that already imports it,
ensuring that the base VM state specification serves as the unique ground truth
for all downstream proofs.

To check all proofs:
\begin{lstlisting}[language=bash]
isabelle build -D proof/
\end{lstlisting}

%% TODO(bob): cite isabelle installation docs (DEVELOPER.md setup section)

The proof status convention used throughout the corpus distinguishes three
levels: a check mark ($\checkmark$) denotes a fully mechanised result with no
outstanding human-audit obligation; a circle ($\circ$) denotes a proof that
depends on an explicit named axiom, where the corresponding C code must be
manually verified to satisfy the axiom; and a warning symbol ($\triangle$)
denotes a use of \texttt{sorry} or a placeholder that must be resolved before
the proof is considered complete. No file in the current corpus uses
\texttt{sorry}.

\section{Base Definitions}
\label{vol1:sec:proofs:base}

%% SOURCE: proof/StarForth_Base.thy

\texttt{StarForth\_Base.thy} is the ground-truth specification of the StarForth
VM state. It defines the \texttt{vm\_state} record type with all fields
corresponding one-to-one to \texttt{include/vm.h}: data stack, return stack,
memory, dictionary, physics substate (rolling window, heat, decay slope,
pipeline metrics, heartbeat), and lock fields. It defines the fundamental
predicates \texttt{wf\_vm} (well-formedness), \texttt{ds\_full} (data stack at
capacity), and \texttt{rs\_full} (return stack at capacity), which must be
audited against the corresponding C guard expressions in \texttt{vm.c} and
\texttt{word\_source/}.

The authority protocol is explicit: C code is written to match this theory,
not the other way around. Discrepancies between this theory and the C source
are treated as bugs in the C code. Every field and predicate carries a
\texttt{CODE-MUST-MATCH} annotation identifying the corresponding C
declaration. \texttt{StarForth\_Base.thy} also imports
\texttt{Main} from the Isabelle/HOL standard library and exports definitions
to all downstream theories.

\section{Physics Loop Proofs}
\label{vol1:sec:proofs:loops}

Seven theory files establish the formal properties of the physics feedback
loops described in Chapter~\ref{vol1:chap:physics}.

\subsection{Loop \#1: Execution Heat}

\textbf{Theory:} \texttt{StarForth\_Loop1\_Heat.thy}.
Mirrors: \texttt{src/dictionary\_heat\_optimization.c}.

The theory proves five properties of the heat lifecycle: heat is non-negative
throughout its lifecycle; heat increment strictly increases \texttt{de\_heat};
heat is bounded above by a \texttt{MAX\_HEAT} constant (preventing integer
overflow in the C field); \texttt{WORD\_FROZEN} words are immune to decay
steps (their heat field is not modified by the decay sweep); and
\texttt{WORD\_PINNED} words hold a heat value of at least~1 after any decay
step. All five results are fully mechanised.

\subsection{Loop \#2: Rolling Window Bounded Growth}

\textbf{Theory:} \texttt{StarForth\_Loop2\_Window.thy}.
Mirrors: \texttt{src/rolling\_window\_of\_truth.c}.

The theory models two window-size quantities: \texttt{rw\_eff\_window} (the
effective window size, which is adaptive and bounded to
$[\texttt{ADAPTIVE\_MIN}, \texttt{ROLLING\_WINDOW\_SIZE}]$) and
\texttt{rw\_act\_window} (the actual fill level of the ring buffer, which grows
monotonically until it saturates at \texttt{ROLLING\_WINDOW\_SIZE}). The
central result establishes that the effective window size can never escape its
bounds regardless of how many shrink or grow operations the inference engine
applies. This prevents the window from collapsing to zero or exceeding the
buffer capacity.

\subsection{Loop \#3: Decay Convergence}

\textbf{Theory:} \texttt{StarForth\_Loop3\_Decay.thy}.
Imports: \texttt{StarForth\_Q48\_16}, \texttt{StarForth\_Base}.
Mirrors: \texttt{src/vm\_time.c}, \texttt{include/vm.h}.

The theory proves that the decay operation is non-negative (heat after decay is
$\max(0, \ldots)$, so it cannot go negative), monotonically decreasing
(repeated decay sweeps never increase heat), and that \texttt{WORD\_FROZEN}
words are fully immune to decay steps. The decay constant
\texttt{DECAY\_RATE\_PER\_US\_Q16} and the adaptive slope
\texttt{decay\_slope\_q48} are both modelled as \Qtype{} quantities using the
fixed-point theory.

\subsection{Loop \#4: Pipelining Metrics Safety}

\textbf{Theory:} \texttt{StarForth\_Loop4\_Pipeline.thy}.
Imports: \texttt{StarForth\_Q48\_16}, \texttt{StarForth\_Base}.
Mirrors: \texttt{src/physics\_pipelining\_metrics.c}.

The theory establishes safety properties for the transition-probability
computation and prefetch machinery: transition counts are non-negative;
the probability formula produces a result in $[0, 1]$ (in \Qtype{} fixed
point); and prefetch-hit/miss accounting preserves the invariant that
\texttt{prefetch\_hits} never exceeds \texttt{prefetch\_attempts}.

\subsection{Loop \#5: Window Width Inference Termination}

\textbf{Theory:} \texttt{StarForth\_Loop5\_WinInf.thy}.
Imports: \texttt{StarForth\_Q48\_16}, \texttt{StarForth\_Loop2\_Window}.
Mirrors: \texttt{src/inference\_engine.c}.

The theory proves that the binary-chop window suggestion always produces a
value within $[\texttt{ADAPTIVE\_MIN\_WINDOW\_SIZE}, \texttt{ROLLING\_WINDOW\_SIZE}]$
before the result is written to the VM, so the effective window size invariant
from Loop~\#2 is preserved by inference updates. The statistical correctness
of the Levene F-test is taken as an explicit named axiom
(\texttt{inference\_window\_clamped\_valid}), carrying a documented audit
obligation against \texttt{run\_inference()} in \texttt{src/inference\_engine.c}.
This is the only axiom in the corpus that depends on a mathematical result
external to the HOL standard library; it is documented as such and does not
use \texttt{sorry}.

\subsection{Loop \#6: Decay Slope Inference Soundness}

\textbf{Theory:} \texttt{StarForth\_Loop6\_DecayInf.thy}.
Imports: \texttt{StarForth\_Q48\_16}, \texttt{StarForth\_Loop3\_Decay}.
Mirrors: \texttt{src/inference\_engine.c}.

The theory proves that the OLS (ordinary least-squares) regression result is
always clamped to $[\texttt{DECAY\_SLOPE\_MIN}, \texttt{DECAY\_SLOPE\_MAX}]$
before being written to the VM state, so the decay slope remains well-formed
regardless of the numerical output of the regression. Unlike Loop~\#5, no
statistical axiom is required; all invariants follow from the clamping alone.
The theory also proves that no update occurs when \texttt{io\_early\_exited}
is set (the ANOVA early-exit guard).

\subsection{Loop \#7: Heartrate Adaptive Stability}

\textbf{Theory:} \texttt{StarForth\_Loop7\_Heartrate.thy}.
Mirrors: \texttt{src/vm\_time.c}, \texttt{include/vm.h}.

The theory proves that \texttt{tick\_target\_ns} is always strictly greater
than zero, remains within $[\texttt{TICK\_MIN\_NS}, \texttt{TICK\_MAX\_NS}]$
under all adjustment operations (both lengthening and shortening the period),
and that \texttt{hb\_tick\_count} increases monotonically---meaning the tick
counter never wraps backward or resets unexpectedly during normal operation.

\section{Word-Category Proofs}
\label{vol1:sec:proofs:words}

%% SOURCE: proof/StarForth_{Arithmetic,Stack,Logical,Memory,Return_Stack,Q48_16}_Words.thy

Six theory files verify the FORTH-79 word implementations against the base
VM state model.

\begin{table}[ht]
\centering
\small
\begin{tabular}{lll}
\toprule
Theory file & Word category & Principal result \\
\midrule
\texttt{StarForth\_Stack\_Words.thy}
  & Stack manipulation (\texttt{DUP}, \texttt{DROP}, \texttt{SWAP}, \texttt{ROT}, \ldots)
  & Stack-depth arithmetic; error on underflow \\
\texttt{StarForth\_Arithmetic\_Words.thy}
  & Arithmetic (\texttt{+}, \texttt{-}, \texttt{*}, \texttt{/}, \texttt{MOD}, \ldots)
  & Binary-op lifting; HOL int (unbounded) semantics \\
\texttt{StarForth\_Logical\_Words.thy}
  & Logic and comparison (\texttt{AND}, \texttt{OR}, \texttt{XOR}, \texttt{INVERT}, \ldots)
  & FORTH-79 boolean convention: TRUE~=~$-1$, FALSE~=~$0$ \\
\texttt{StarForth\_Return\_Stack\_Words.thy}
  & Return stack (\texttt{>R}, \texttt{R>}, \texttt{R@}, \texttt{RDROP}, \ldots)
  & Bidirectional stack transfer; capacity bounds \\
\texttt{StarForth\_Memory\_Words.thy}
  & Memory (\texttt{@}, \texttt{!}, \texttt{C@}, \texttt{C!}, \texttt{MOVE}, \texttt{FILL})
  & Read-after-write correctness; absence of spurious mutation \\
\texttt{StarForth\_Q48\_16.thy}
  & Q48.16 fixed-point arithmetic
  & HOL-Word 64-bit model; wrapping matches C \texttt{uint64\_t} \\
\bottomrule
\end{tabular}
\caption{Word-category proof inventory.}
\label{vol1:tab:proofs:words}
\end{table}

All word-category theories import \texttt{StarForth\_Base} and use HOL
integers (arbitrary precision) as the semantic domain, with the understanding
that proofs hold in C provided no intermediate value overflows 64-bit signed
range. \texttt{StarForth\_Q48\_16.thy} uses \texttt{HOL-Library.Word} to model
the 64-bit fixed-point type with the same wrapping arithmetic as C
\texttt{uint64\_t}, closing the gap that an earlier natural-number model left
open.

\section{Concurrency and Mutual Exclusion}
\label{vol1:sec:proofs:concurrent}

%% SOURCE: proof/StarForth_Concurrent.thy, proof/StarForth_Mutex.thy,
%%         proof/StarForth_Transition.thy

Three theory files establish the concurrency model and its safety properties.

\texttt{StarForth\_Transition.thy} (imports \texttt{StarForth\_Mutex})
defines the Labeled Transition System governing the interaction between the
execution thread (\texttt{ExecThread}) and the heartbeat thread
(\texttt{HeartbeatThread}). It introduces the quotient relation
\texttt{exec\_equiv} ($\simeq$) and establishes two explicit named axioms:
\texttt{heartbeat\_exec\_neutral} (the heartbeat step is the identity in
\texttt{vm\_state}/$\simeq$; audited against \texttt{src/vm\_time.c}) and
\texttt{word\_physics\_transparent} (word execution is a congruence law for
$\simeq$; audited against the Level-1 word theories). The former eight
field-level axioms present in earlier versions of the theory are now proved
from these two axioms via the \texttt{exec\_equiv} quotient.

\texttt{StarForth\_Mutex.thy} (imports \texttt{StarForth\_Base}) models the
two VM locks---\texttt{tuning\_lock} (guards physics-loop parameter updates)
and \texttt{dict\_lock} (guards dictionary writes)---as first-class
\texttt{vm\_state} fields of type \texttt{lock\_state}
(\texttt{LockFree}~$|$~\texttt{LockHeld nat}). The mutual-exclusion safety
property is proved: at most one thread holds a given lock at any time. Since
StarForth has exactly two threads (execution and heartbeat), thread identifiers
are $0$ (\texttt{ExecThread}) and $1$ (\texttt{HeartbeatThread}).

\texttt{StarForth\_Concurrent.thy} (imports \texttt{StarForth\_Transition},
\texttt{StarForth\_Loop7\_Heartrate}, \texttt{StarForth\_Loop3\_Decay})
proves non-interference: the data-stack result of executing any word is the
same regardless of how many heartbeat steps have interleaved, formalized as
the \texttt{heartbeat\_noninterference} theorem:

\begin{lstlisting}[language=isabelle]
theorem heartbeat_noninterference:
  "\<forall>n k vm.
     data_stack (word_table (heartbeat_step ^^ k $ vm) n
                             (heartbeat_step ^^ k $ vm))
     = data_stack (word_table vm n vm)"
\end{lstlisting}

This result, which is sorry-free and proved from the two axioms in
\texttt{StarForth\_Transition}, establishes that arbitrary word sequences are
independent of heartbeat timing---the adaptive physics machinery does not
disturb the observable stack behavior of executing programs.

\section{Overall Correctness}
\label{vol1:sec:proofs:correctness}

%% SOURCE: proof/StarForth_Correctness.thy

\texttt{StarForth\_Correctness.thy} is the top-level theory that imports all
twelve subordinate theories and assembles the totality result. It is
sorry-free, with an explicit inventory of two axioms (both inherited from
\texttt{StarForth\_Transition}) and no others. The top-level correctness
theorem binds the word-category correctness results, the loop invariants, and
the concurrency non-interference result into a single statement asserting that
the StarForth VM satisfies its specification across all defined operating
conditions.

\section{Proof Inventory}
\label{vol1:sec:proofs:inventory}

Table~\ref{vol1:tab:proofs:all} lists all 24 theory files with a one-line
description of the principal result or role of each.

\begin{table}[ht]
\centering
\small
\begin{tabular}{ll}
\toprule
Theory file & Principal result or role \\
\midrule
\texttt{StarForth\_Base.thy}
  & Ground-truth VM state type; base predicates (\texttt{wf\_vm}, etc.) \\
\texttt{StarForth\_Q48\_16.thy}
  & Q48.16 fixed-point: HOL-Word 64-bit model matching C \texttt{uint64\_t} \\
\texttt{StarForth\_Stack\_Words.thy}
  & Stack-manipulation words: depth arithmetic, underflow error \\
\texttt{StarForth\_Arithmetic\_Words.thy}
  & Arithmetic words: binary-op lifting, HOL-int semantics \\
\texttt{StarForth\_Logical\_Words.thy}
  & Logical/comparison words: FORTH-79 boolean convention \\
\texttt{StarForth\_Return\_Stack\_Words.thy}
  & Return-stack words: bidirectional transfer, capacity bounds \\
\texttt{StarForth\_Memory\_Words.thy}
  & Memory words: read-after-write correctness, no spurious mutation \\
\texttt{StarForth\_Loop1\_Heat.thy}
  & Loop \#1: heat non-negativity, MAX\_HEAT bound, FROZEN/PINNED semantics \\
\texttt{StarForth\_Loop2\_Window.thy}
  & Loop \#2: effective window bounded in [ADAPTIVE\_MIN, ROLLING\_WINDOW\_SIZE] \\
\texttt{StarForth\_Loop3\_Decay.thy}
  & Loop \#3: decay non-negativity, monotone decrease, FROZEN immunity \\
\texttt{StarForth\_Loop4\_Pipeline.thy}
  & Loop \#4: transition-probability in $[0,1]$, hit/attempt accounting safety \\
\texttt{StarForth\_Loop5\_WinInf.thy}
  & Loop \#5: inference output clamped in bounds; Levene axiom audited \\
\texttt{StarForth\_Loop6\_DecayInf.thy}
  & Loop \#6: OLS slope clamped, no update on early-exit (sorry-free) \\
\texttt{StarForth\_Loop7\_Heartrate.thy}
  & Loop \#7: tick period in $[\texttt{TICK\_MIN\_NS}, \texttt{TICK\_MAX\_NS}]$, monotone count \\
\texttt{StarForth\_Mutex.thy}
  & Mutual exclusion: at most one thread holds \texttt{tuning\_lock}/\texttt{dict\_lock} \\
\texttt{StarForth\_Transition.thy}
  & Labeled transition system; \texttt{exec\_equiv} quotient; 2 explicit axioms \\
\texttt{StarForth\_Concurrent.thy}
  & Non-interference: heartbeat steps do not affect observable stack results \\
\texttt{StarForth\_Correctness.thy}
  & Top-level totality theorem; imports all 12 theories; sorry-free \\
\texttt{ACL\_Pin\_Monotone.thy}
  & \texttt{acl\_pinned} is set-only: once set to~1, never cleared to~0 \\
\texttt{ACL\_Inherit\_Clears\_Pin.thy}
  & \texttt{ACL-INHERIT} always clears \texttt{acl\_pinned} on the destination entry \\
\texttt{ACL\_TTL\_Bounded.thy}
  & TTL output is $\ge \texttt{ACL\_BASE\_TTL}$ and $\le \texttt{ACL\_MAX\_TTL}$; safe decrement \\
\texttt{ACL\_Emergency\_Bypass.thy}
  & \texttt{emergency\_console=1} or \texttt{zuse\_session=1} bypasses all ACL checks \\
\texttt{ACL\_No\_Escalation.thy}
  & A child word cannot acquire \texttt{acl\_pinned=1} via \texttt{ACL-INHERIT} alone \\
\texttt{ROOT}
  & Isabelle session manifest (not a theory; declares import order) \\
\bottomrule
\end{tabular}
\caption{All 24 entries in the \texttt{proof/} corpus (23 \texttt{.thy} files + \texttt{ROOT}).}
\label{vol1:tab:proofs:all}
\end{table}

\section{ACL System Proofs}
\label{vol1:sec:proofs:acl}

%% SOURCE: proof/ACL_Pin_Monotone.thy, ACL_Inherit_Clears_Pin.thy,
%%         ACL_TTL_Bounded.thy, ACL_Emergency_Bypass.thy, ACL_No_Escalation.thy
%%         .claude/CLAUDE.md §Word-Level ACL System

Five theory files verify the word-level ACL system implemented in
\texttt{src/word\_source/acl\_words.c} and \texttt{capsules/ACL.4th}.
All five import \texttt{StarForth\_Base} or a theory that imports it; all are
sorry-free.

\textbf{\texttt{ACL\_Pin\_Monotone.thy}} proves that the \texttt{acl\_pinned}
field in \texttt{DictEntry} is a set-only (monotone) field: once written to~1
by \texttt{ACL-PIN}, no subsequent operation---\texttt{ACL-MODE!},
\texttt{ACL-TTL!}, or \texttt{ACL-ALLOW!}---can clear it to~0. The
\texttt{CODE-MUST-MATCH} annotation lists four C functions that must never
write \texttt{acl\_pinned=0} after it has been set. This property ensures that
pinned words retain their immutability guarantee for the lifetime of the VM
session.

\textbf{\texttt{ACL\_Inherit\_Clears\_Pin.thy}} proves that
\texttt{ACL-INHERIT} always clears \texttt{acl\_pinned} on the
\emph{destination} entry, even when the source entry is pinned. The
\texttt{acl\_inherit\_entry()} C function mirrors this by setting
\texttt{dst->acl\_pinned = 0} unconditionally. This prevents inherited ACL
from accidentally propagating the immutability of a privileged parent word to
a child.

\textbf{\texttt{ACL\_TTL\_Bounded.thy}} proves three properties of the
adaptive TTL mechanism: the output of \texttt{ACL-TTL-COMPUTE} is always at
least \texttt{ACL\_BASE\_TTL} (minimum service guarantee); it is always at
most \texttt{ACL\_MAX\_TTL} (bounded cost amortization); and the hot-path TTL
decrement is safe (TTL cannot go below~0 when the pre-condition holds).
Additionally, STRICT mode holds TTL at~0 after every \texttt{ACL-RECHECK}.

\textbf{\texttt{ACL\_Emergency\_Bypass.thy}} proves that when either
\texttt{emergency\_console=1} (fault handler active) or
\texttt{zuse\_session=1} (superuser authenticated) is set in the VM flags, all
ACL checks are bypassed. The two conditions are independent: the emergency
console flag is set only by C code (\texttt{acl\_recheck}, \texttt{sk\_repl})
during error recovery; the superuser flag is set only by the
\texttt{ZUSE-AUTHENTICATE} C primitive. The theory proves that neither
condition can be established from within the ACL policy layer itself.

\textbf{\texttt{ACL\_No\_Escalation.thy}} (imports
\texttt{ACL\_Inherit\_Clears\_Pin}, \texttt{ACL\_TTL\_Bounded},
\texttt{ACL\_Emergency\_Bypass}) proves the central security property: a child
word cannot acquire \texttt{acl\_pinned=1} through \texttt{ACL-INHERIT} alone.
Inheriting from any source word---pinned or not---always leaves the destination
un-pinned. A word can become pinned only through an explicit \texttt{ACL-PIN}
call, which requires intentional action from a trusted caller. This prevents
capsule words from upgrading themselves to kernel-level immutability by
inheriting from a privileged parent.
