% FORTH Neuronet: A Conservation-Law-Governed Cognitive Mesh Architecture
% Executive Summary — Starship Ananke Foundation
% Robert Allan James — June 2026

\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{parskip}
\usepackage{enumitem}

\title{\textbf{FORTH Neuronet}\\
       \large A Conservation-Law-Governed Cognitive Mesh Architecture\\[0.5em]
       \normalsize Executive Summary --- Starship Ananke Foundation}
\author{Robert Allan James}
\date{June 2026}

\begin{document}

\maketitle

\begin{abstract}
We describe a novel computational architecture emerging from the intersection
of three independently validated components: the StarshipOS capsule lifecycle
substrate, the FORTH execution model, and conservation-law-governed state
management via $K \equiv 1.0$. Together these components suggest a tractable
path toward modular, self-modifying, runtime-plastic cognitive mesh
computing---distinct from and complementary to current transformer-based AI
architectures. Prior art appears absent. The substrate was built first. The
conservation law was proven first. The architecture that emerges from them is
not accidental.
\end{abstract}

\tableofcontents
\newpage

% ============================================================================
\section{The Core Observation}

The FORTH execution model is an unrecognized neural computation substrate.
This is not metaphor. It is a technically grounded structural observation.

\begin{table}[h]
\centering
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{FORTH construct} & \textbf{Neural analog} \\
\midrule
Parameter stack          & Membrane potential accumulator \\
Word                     & Deterministic threshold function \\
Dictionary lookup        & Weighted connectivity \\
Return stack             & Temporal and refractory state \\
Word redefinability      & Runtime synaptic plasticity \\
Word composition         & Connection topology \\
\bottomrule
\end{tabular}
\caption{Structural correspondence between FORTH primitives and neural
         computation. Each mapping is exact, not analogical.}
\end{table}

No prior research appears to have formalized this observation.
Chuck Moore's colorFORTH approached computational minimalism from the
opposite direction---toward irreducible simplicity rather than toward
emergent complexity from constrained simple structure.

% ============================================================================
\section{The Danger and Its Resolution}

A self-modifying FORTH neuronet without constraint is formally unpredictable.
Dictionary rewriting cascades produce execution paths that diverge without
bound. This is the reason the idea has remained unexplored---it appears
ungovernable.

$K \equiv 1.0$ resolves this.

The StarshipOS Steady-State Machine conservation substrate provides the
thermodynamic constraint that makes runtime plasticity tractable:

\begin{itemize}
  \item Stack energy is conserved across all word invocations
  \item Reasoning lineage is preserved via the ACL Rolling Window of Truth (RWT)
  \item The network may rewire itself within a provably bounded state space
\end{itemize}

Let $\mathcal{W}$ be the dictionary state space and $\phi: \mathcal{W}
\rightarrow \mathcal{W}$ be any self-modifying transition. The conservation
law asserts:

\begin{equation}
  K(\phi(\mathcal{W})) \equiv K(\mathcal{W}) = 1.0
\end{equation}

where $K$ is the steady-state convergence invariant proven experimentally
with $p = 1.000$ across 90 independent runs on three ISAs. The conservation
law is the leash. LithosAnanke is the containment vessel.

% ============================================================================
\section{The Architectural Hierarchy}

The granularity of computation maps cleanly to biological hierarchy without
simulating biology.

\begin{table}[h]
\centering
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Biological} & \textbf{FORTH Neuronet} & \textbf{Implementation} \\
\midrule
Neuron              & Stack cell              & \texttt{cell\_t} \\
Synapse             & Word invocation edge    & Dictionary link \\
Cortical column     & StarForth VM instance   & \texttt{VM} struct \\
White matter tract  & Gefyra inter-VM fabric  & (planned) \\
Brain region        & Capsule cluster         & \texttt{CapsuleDesc} \\
\bottomrule
\end{tabular}
\caption{Mapping between biological neural hierarchy and FORTH Neuronet
         components. Column three shows the existing StarForth implementation
         artifact at each level.}
\end{table}

The VM is not wasteful as a neuron---it is correctly sized as a cortical
column. Each VM emits not a binary spike but a rich state vector into the
Gefyra fabric. The mesh receives richer signal than biological tissue.
Intelligence emerges from topology and conservation, not from individual
node complexity.

% ============================================================================
\section{The LLM Capsule}

Current large language models are monolithic. Context degrades entropically.
There is no conservation law governing what persists across the rolling
attention window. Signal and noise fall off the edge with equal weight.

A large language model (e.g., Llama) deployed as a LithosAnanke capsule
within a FORTH neuronet mesh inverts this entirely:

\begin{enumerate}
  \item \textbf{BIRTH} on reasoning task arrival
  \item \textbf{FORTH neuronet as cognitive substrate}---the LLM reasons
        \emph{through} the network, not independently of it
  \item \textbf{RWT maintains reasoning lineage} as a security and
        epistemic invariant
  \item \textbf{$K \equiv 1.0$ governs conservation} of established
        truth across the mesh
  \item \textbf{Parent orchestrator holds the Rolling Window of Truth}---not
        conversation history but crystallized verified facts
  \item \textbf{Specialist baby LLM capsules} spin up, execute narrow
        focused tasks with full fresh context, report back, \textbf{DIE}
\end{enumerate}

No context pollution. No recursive spiral. No entropic degradation.

The LLM is no longer the intelligence. It is one capsule in a larger
cognitive architecture with a provable conservation law underneath.

% ============================================================================
\section{Differentiation from Prior Work}

\begin{table}[h]
\centering
\begin{tabular}{@{}lp{9cm}@{}}
\toprule
\textbf{Approach} & \textbf{Why it differs} \\
\midrule
RAG &
  Bolts external memory onto a monolithic model. No conservation substrate.
  Context window still degrades. \\
Mixture of Experts &
  Routes tokens within a single model, not across a lifecycle-governed
  capsule mesh. No formal invariant. \\
Neuromorphic computing &
  Hardware simulation of biological spiking neurons. This is a
  software-layer structural observation, not a hardware emulation. \\
Agent frameworks &
  LangChain, AutoGen, and similar lack a conservation law substrate and
  formal capsule lifecycle governance. Stateful sessions accumulate noise. \\
\bottomrule
\end{tabular}
\caption{Differentiation from adjacent approaches. None of the listed
         approaches combines conservation-law substrate with formal capsule
         lifecycle governance.}
\end{table}

% ============================================================================
\section{Prior Art}

Apparently none. The specific intersection---FORTH execution model as neural
substrate, capsule lifecycle as cortical column, conservation law as
thermodynamic constraint on a cognitive mesh---does not appear in the
literature. A thorough search is warranted before formal publication, but
the combination is novel by construction: the three components were
developed independently for unrelated reasons and their convergence was
not anticipated.

% ============================================================================
\section{Next Steps}

\begin{enumerate}
  \item Literature search confirming prior art gap
  \item Formal specification of FORTH word as threshold function $\theta: S^n \rightarrow S$
  \item Stack energy conservation proof under $K \equiv 1.0$
  \item Prototype neuronet in StarForth: minimal mesh, three capsule VMs,
        Gefyra fabric
  \item SSRN preprint: \emph{FORTH Neuronet: Conservation-Law-Governed
        Cognitive Mesh Architecture}
  \item Patent evaluation with current counsel
\end{enumerate}

% ============================================================================
\section{Relationship to Existing Work}

All substrate components are implemented and experimentally validated:

\begin{description}[style=nextline]
\item[$K \equiv 1.0$ proof]
  \texttt{docs/02-experiments/james-law/protocol.md}

\item[Capsule lifecycle]
  \texttt{src/starkernel/capsule/} --- birth, validation, content-hash
  identity, parity logging

\item[Steady-State Machine]
  \texttt{src/ssm\_jacquard.c};
  \texttt{docs/03-architecture/physics-engine/steady-state-machine.md}

\item[ACL-RWT (reasoning lineage invariant)]
  \texttt{capsules/ACL.4th};
  \texttt{docs/03-architecture/word-acl/DESIGN.md}

\item[Gefyra fabric]
  Planned; inter-VM transport layer

\item[Prior SSRN paper]
  \texttt{papers/James\_Steady-State\_Convergence\_Adaptive\_Runtime.pdf}
\end{description}

The substrate was built first. The conservation law was proven first.
The architecture that emerges from them is not accidental.

\end{document}
