%% SCRAP: scratch/src/development/CONTRIBUTING
%% SOURCE: docs/working/scratch/src/development/CONTRIBUTING.adoc
%% STATUS: WORKING
%% FITS: dev-guide/ — contains the current contribution workflow; assess whether the
%%   placeholder fields (project name, CoC contact) have been populated before promoting.
%% EDITORIAL: lifted — prose rewritten to press voice

\section{Contributing to StarshipOS}

Contributions are welcome. This section documents the workflow, standards, and
conventions required for a pull request to be accepted.

\subsection{Quick Summary}

Fork the repository, create a focused branch, open small pull requests, follow
Conventional Commits, run \texttt{make fmt lint test} before pushing, include
reproduction steps, documentation, and tests, and treat collaborators with directness
and respect.

\subsection{Code of Conduct}

The project follows the Contributor Covenant. Violations may result in removal from
participation. %% TODO(bob): supply the incident-report contact address.

\subsection{Getting Started}

\paragraph{Prerequisites.}
Linux, macOS, or WSL2; \texttt{git}, \texttt{make}; the C99 toolchain described in the
developer guide; and \texttt{pre-commit} for hook management
(\texttt{pipx install pre-commit \&\& pre-commit install}).

\paragraph{Setup.}
\begin{lstlisting}[language=bash]
git clone https://github.com/<you>/StarForth.git
cd StarForth
make build
make test
\end{lstlisting}

\paragraph{Local development loop.}
\begin{lstlisting}[language=bash]
make fmt      # auto-format
make lint     # static analysis
make test     # unit tests
make itest    # integration tests (if available)
\end{lstlisting}

\subsection{Branching and Commit Style}

Branch from \texttt{main}. Name branches \texttt{feat/<slug>}, \texttt{fix/<slug>},
\texttt{docs/<slug>}, or \texttt{chore/<slug>}. Write commits in the Conventional
Commits format with imperative mood, subject line at most 72 characters, and issue
references (\texttt{Fixes \#123} or \texttt{Refs \#456}). Sign every commit with a
Developer Certificate of Origin \texttt{Signed-off-by} trailer.

Examples:
\begin{lstlisting}
feat(runtime): add cgroups v2 isolation flag
fix(scheduler): prevent null deref on empty queue
docs: clarify quickstart on macOS
\end{lstlisting}

\subsection{Pull Request Standards}

Keep pull requests under approximately 300 lines changed. Each PR must: link to an
issue or explain its rationale; add or update tests; update documentation; pass
\texttt{make fmt lint test} locally; address backwards compatibility; and measure
performance impact where relevant. Draft PRs are welcome early.

Open an issue first for large or controversial changes.

\subsection{Architecture Decision Records}

Add an ADR (\texttt{docs/adrs/ADR-XXXX-title.md}) for any non-trivial design choice.

%% TODO(bob): verify that the ADR directory exists and confirm the numbering convention.
