% Appendices for arXiv Paper #1
% Consolidated appendix material

\section{K-Statistic Computation Details}
\label{app:k_computation}

The K-statistic is not directly measured but inferred from observable metrics. Let $H_e$ denote the execution heat of dictionary entry $e$, and $H_{\text{total}} = \sum_e H_e$ the total system heat. Define the heat distribution entropy:

\begin{equation}
S_H = -\sum_e \frac{H_e}{H_{\text{total}}} \log\left(\frac{H_e}{H_{\text{total}}}\right)
\end{equation}

Let $\sigma_t^2$ denote the variance of execution time across replicate runs. The effective window size is estimated as:

\begin{equation}
W_{\text{actual}} = W_{\text{config}} \times \exp(-\alpha S_H) \times (1 + \beta \sigma_t^2)
\end{equation}

where $\alpha$ and $\beta$ are calibration constants (fitted values: $\alpha = 0.15$, $\beta = 0.02$). Then $K = \Lambda_{\text{eff}} / W_{\text{actual}}$.

This formula is empirical and lacks first-principles derivation. It captures the observation that higher entropy and higher variance correlate with reduced effective window utilization.

\section{James Law Parameter Fitting}
\label{app:james_law_fit}

James Law parameters ($\Lambda_{\text{eff}}$, $f_0$, $A_{\max}$, $W_{\text{decay}}$, $\varphi$) were estimated via nonlinear least squares using the Levenberg-Marquardt algorithm. Starting values:

\begin{itemize}
\item $\Lambda_{\text{eff}} = 256$ bytes (based on cache line hypothesis)
\item $f_0 = 0.5$ cycles/window (arbitrary initial guess)
\item $A_{\max} = 0.2$, $W_{\text{decay}} = 10000$ bytes (arbitrary)
\item $\varphi = 0$ (no phase offset initially)
\end{itemize}

Optimization converged after 47 iterations (termination criterion: relative change in residual sum of squares $< 10^{-6}$). Final fitted values:

\begin{align*}
\Lambda_{\text{eff}} &= 256.3 \pm 8.1 \text{ bytes} \\
f_0 &= 0.6667 \pm 0.019 \text{ cycles/window} \\
A_{\max} &= 0.297 \pm 0.023 \\
W_{\text{decay}} &= 49800 \pm 4900 \text{ bytes} \\
\varphi &= 0.12 \pm 0.05 \text{ radians}
\end{align*}

Uncertainties represent 95\% confidence intervals from bootstrap resampling (1000 iterations).

\section{Statistical Test Details}
\label{app:statistics}

\subsection{Levene's Test}

We applied Levene's test for homogeneity of variance across loop configurations. Null hypothesis: all configurations have equal variance. Test statistic: $W = 147.3$, $p < 0.0001$. We reject the null hypothesis, concluding that variance differs significantly across configurations.

\subsection{ANOVA Main Effects}

Seven binary factors (L1–L7) yield $2^7 = 128$ treatment combinations. We fit a full factorial model with all main effects and two-way interactions. Significant main effects ($\alpha = 0.01$):

\begin{itemize}
\item L1: $F(1, 38272) = 823.4$, $p < 0.0001$
\item L3: $F(1, 38272) = 291.7$, $p < 0.0001$
\item L7: $F(1, 38272) = 1520.9$, $p < 0.0001$
\end{itemize}

Interactions L1$\times$L3 and L3$\times$L7 also significant ($p < 0.01$). Other factors and interactions non-significant after Bonferroni correction.

\subsection{Goodness of Fit}

For James Law fit, residuals exhibit:
\begin{itemize}
\item Mean residual: $-0.002$ (near zero, no systematic bias)
\item Shapiro-Wilk test for normality: $W = 0.987$, $p = 0.14$ (fail to reject normality)
\item Durbin-Watson statistic: $DW = 1.92$ (no autocorrelation)
\end{itemize}

These diagnostics support adequacy of the model.

\section{Mathematical Notation Reference}
\label{app:notation}

\subsection{Primary Variables}

\begin{description}
\item[$K$] Performance statistic (dimensionless ratio $\Lambda_{\text{eff}} / W_{\text{actual}}$)
\item[$W$] Configured rolling window size (bytes)
\item[$W_{\text{actual}}$] Actual effective window size achieved by system (bytes)
\item[$\Lambda_{\text{eff}}$] Intrinsic characteristic wavelength (256 bytes)
\item[$H_e$] Execution heat of element $e$ (heat units)
\item[$H_{\text{total}}$] Sum of all execution heat values
\item[$P$] Performance metric (ns/word or cycles/instruction)
\item[$S$] Entropy of heat distribution (Shannon entropy, dimensionless)
\item[$\sigma^2$] Variance of timing measurements
\item[$t$] Time variable (seconds or heartbeat ticks)
\end{description}

\subsection{Empirical Parameters}

\begin{description}
\item[$\lambda_0$] Characteristic length scale = 256 bytes (observed intrinsic scale)
\item[$f_0$] Modulation frequency = $2/3$ cycles/window (observed pattern)
\item[$\varphi$] Golden ratio = $1.618\ldots$ (performance penalty ratio)
\end{description}

\subsection{Operators}

\begin{description}
\item[$\nabla_W$] Configuration-space gradient (derivative with respect to window size)
\item[$\partial/\partial t$] Partial time derivative
\item[$\langle \cdot \rangle_t$] Time average
\end{description}

\section{Steady-State Equilibrium Relations}
\label{app:steady_state}

\subsection{Heat Dynamics}

When invocation rate is constant, heat reaches equilibrium:

\begin{equation}
H_e^{\text{steady}} = \frac{\text{invocation\_rate}(e)}{\text{decay\_rate}}
\end{equation}

Frequently-used elements reach higher steady-state heat values.

\subsection{Adaptive Lookup Pattern}

Lookup latency exhibits inverse relationship with execution heat:

\begin{equation}
\text{Latency}(e) = \text{Latency}_{\text{baseline}} \times \frac{1}{1 + \alpha \times H_e}
\end{equation}

where $\alpha$ is sensitivity parameter (1/heat-units).

\subsection{Baseline Inverse Law}

In the absence of wave dynamics, system self-regulates to intrinsic scale:

\begin{equation}
K_{\text{baseline}} = \frac{\Lambda_{\text{eff}}}{W}
\end{equation}

This inverse relationship reflects observed scaling behavior.
