%% SCRAP: scratch/GOVERNANCE_EXPORT_NOTES
%% SOURCE: docs/working/scratch/GOVERNANCE_EXPORT_NOTES.adoc
%% STATUS: HISTORICAL
%% FITS: none
%% EDITORIAL: lifted — prose rewritten to press voice

\section{Governance Export Notes --- Physics Phase 1}

This checklist records which Phase 1 artefacts require mirroring into the
\texttt{StarForth-Governance} repository.

\begin{center}
\begin{tabular}{p{3.5cm}p{4.5cm}p{5cm}}
\toprule
Artefact & Location (VM repo) & Governance action \\
\midrule
\texttt{physics\_runtime.h} & \texttt{include/physics\_runtime.h} &
  Copy verbatim; annotate protocol version and magic constants. \\
HOLA protocol memo & \texttt{docs/src/internal/HOLA\_PROTOCOL.adoc} &
  Mirror as governance RFC; retain version history. \\
Physics formal models &
  \texttt{Physics\_StateMachine.thy}, \texttt{Physics\_Observation.thy} &
  Import into governance proofs tree for cross-repository review. \\
Scheduler plan updates & \texttt{PHYSICS\_SCHEDULING\_PLAN.adoc} &
  Reference from governance scheduling playbook (Phase 1 appendix). \\
Signal map updates & \texttt{PHYSICS\_SIGNAL\_MAP.adoc} &
  Link from governance telemetry charter. \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Process Notes}

The governance repository holds the source of truth for ABI versions. Any change to
\texttt{HOLA\_PROTOCOL.adoc} or the analytics header must update
\texttt{HOLA\_PROTOCOL\_VERSION} in the governance repository before landing in the VM
repository. Isabelle theories should be bundled with build instructions so that
governance reviewers can replay proofs independently. Export scripts must verify the
analytics heap size and event channel table to keep governance dashboards consistent
with the VM implementation.

Phase 1 governance deliverables live under \texttt{governance/physics/phase1/}.

\subsection{Open Items}

\begin{itemize}
  \item Decide whether the governance repository ships pre-built analyser binaries
        or only the protocol description.
  \item Clarify signature requirements for future ABI bumps (magic change or command
        addition).
  \item Define the approval flow for new HOLA opcodes before Phase 2 experiments begin.
\end{itemize}
