%% SCRAP: scratch/src-index
%% SOURCE: docs/working/scratch/src-index.adoc
%% STATUS: HISTORICAL
%% FITS: none
%% EDITORIAL: lifted — prose rewritten to press voice

\section{Source Documentation Index (October 2025 Snapshot)}

This index recorded the documentation tree as of October 2025, providing a navigation
map across seven major topic areas. The links it contained pointed to documents that have
since been reorganised under \texttt{docs/working/} and \texttt{docs/formal/}.

\subsection{Topic Areas}

\begin{itemize}
  \item \textbf{Getting started.} Quick-start guide, installation guide, and architecture
        deep dive.
  \item \textbf{Build and tooling.} Build options reference, profile-guided optimisation
        guide, init tools, platform build guide, and documentation system.
  \item \textbf{Architecture and internals.} ABI reference, init system, block storage
        guide, messaging system, and word development guide.
  \item \textbf{Platform integration.} Platform abstraction layer, L4Re integration,
        L4Re dictionary allocation, L4Re block I/O endpoints, and Raspberry Pi build.
  \item \textbf{Performance and profiling.} x86\_64 assembly optimisations, ARM64
        assembly strategy, and profiler guide.
  \item \textbf{Testing and quality.} Test suite guide, break-me report, and gap
        analysis.
  \item \textbf{Development and contribution.} Contributing guide, repository guidelines,
        Doxygen style guide, and Doxygen quick reference.
  \item \textbf{Governance and compliance.} Governance framework, FORTH-79 compliance
        matrix, and validation book (all in the companion governance repository).
\end{itemize}

%% TODO(bob): reconcile this topic map against the current formal/ volume structure
%% and update the Index.md accordingly.
