%% SCRAP: archive/legacy/phase-1/Reference/physics_experiment/PEER_REVIEW_SUBMISSION/README
%% SOURCE: docs/working/archive/legacy/phase-1/Reference/physics_experiment/PEER_REVIEW_SUBMISSION/README.md
%% STATUS: HISTORICAL
%% FITS: none
%% EDITORIAL: lifted — prose rewritten to press voice

\section{Peer Review Submission Package --- Navigation Record}

This document records the structure and status of the peer-review submission package
as of its last update (2025-11-07, approximately 75\% complete, two weeks estimated
to submission). It is a navigation artefact; all technical content resides in the
sibling files.

\subsection{Package Structure}

\begin{itemize}
  \item \texttt{01\_MAIN\_PAPER/PEER\_REVIEW\_PAPER\_DRAFT.md} --- 18-page complete paper
        draft. All eight sections written; citations and figures required before
        submission.
  \item \texttt{01\_MAIN\_PAPER/PAPER\_KEY\_CONTRIBUTIONS\_SUMMARY.md} --- Four-page
        visual summary of the four main contributions with anticipated reviewer
        objections and responses.
  \item \texttt{02\_SUPPORTING\_DOCS/} --- Formal verification interpretation, formal
        claims for reviewers, variance analysis summary, and submission venue checklist.
  \item \texttt{03\_EXPERIMENTAL\_DATA/full\_90\_run\_comprehensive/experiment\_results.csv}
        --- 91-row validated dataset (1 header + 90 runs), 33 metrics per run, zero
        corruption.
  \item \texttt{04\_ANALYSIS\_SCRIPTS/} --- \texttt{analyze\_thermal\_hypothesis.py}
        and \texttt{analyze\_variance\_sources.py}.
  \item \texttt{05\_SUPPLEMENTARY/} --- Completion status, OS optimisation guide, and
        paper outline.
\end{itemize}

\subsection{Venue Recommendation at Time of Archival}

Primary target: IEEE Transactions on Software Engineering. Backup options: ACM TOPLAS
(20--25\% acceptance rate) and Journal of Systems Software (35--45\% acceptance rate,
more practical focus).

%% TODO(bob): record the actual submission venue and outcome.
