%% Vol II, Chapter 2 — Capsule System
%% SOURCE: .claude/CLAUDE.md (Capsule System section)
%%         docs/formal/scraps/architecture/MAMA_FORTH_SUPERVISOR_ARCHITECTURE.tex
%%         src/starkernel/capsule/ (6 files)
%%         capsules/ directory (19 .4th files)

\chapter{Capsule System}
\label{vol2:chap:capsules}

A capsule is an immutable, content-addressed VM initialization payload.
The capsule identifier is the XXHash64 hash of the capsule's content;
any mutation of the payload produces a different hash and is therefore
detectable at birth time. Capsules provide the mechanism by which new
FORTH virtual machine instances are initialized in a verifiable and
reproducible manner, whether on a hosted Linux system or within the
LithosAnanke bare-metal kernel.

\section{Capsule Types}
\label{vol2:sec:capsules:types}

%% SOURCE: .claude/CLAUDE.md (Capsule types)

Three capsule types exist, distinguished by a single type-code field
in the capsule descriptor.

\begin{table}[ht]
  \centering
  \caption{Capsule types}
  \label{tab:capsule-types}
  \begin{tabular}{lll}
    \toprule
    Type code & Name & Purpose \\
    \midrule
    \texttt{(m)} & \texttt{MAMA\_INIT} & Exactly one Mama VM initialization capsule \\
    \texttt{(p)} & \texttt{PRODUCTION} & Truth-bearing baby VM initializers \\
    \texttt{(e)} & \texttt{EXPERIMENT} & DoE workload-only initializers \\
    \bottomrule
  \end{tabular}
\end{table}

Exactly one \texttt{MAMA\_INIT} capsule exists per deployment. It initializes
the root VM (Mama Forth) from which all child VMs are subsequently born.
\texttt{PRODUCTION} capsules carry the authoritative personality for
truth-bearing baby VMs. \texttt{EXPERIMENT} capsules supply workload-only
initialization for Design of Experiments runs; they do not carry personality
state and are not used as birth-protocol truth bearers.

\section{Content Addressing and Immutability}
\label{vol2:sec:capsules:hash}

%% SOURCE: .claude/CLAUDE.md (Capsule System section)
%% SOURCE: src/starkernel/hash/xxhash64.c

The capsule identifier is the XXHash64 hash of the capsule binary content,
computed by \texttt{src/starkernel/hash/xxhash64.c}. Because the identifier
is derived from the content, any alteration to the payload --- including
changes to capsule code, CA-key constants, or block layout --- produces a
different hash. The birth protocol validates this hash before allocating a
VM ID, making the content-hash serve as the root-of-trust fingerprint for
each capsule.

Capsule immutability has a further consequence for the ACL subsystem: the
CA-root public-key placeholder embedded in \texttt{ACL.4th} (Block 2066)
is part of the hashed content. Replacing the placeholder with a real
Ed25519 key necessarily changes the capsule hash, and any such change is
detectable by the birth protocol.

\section{Birth Protocol}
\label{vol2:sec:capsules:birth}

%% SOURCE: .claude/CLAUDE.md (Birth protocol section)

The capsule birth protocol executes the following steps in order:

\begin{enumerate}
  \item Locate the capsule by name in the capsule directory.
  \item Validate the content hash (XXHash64); reject the capsule if the
        computed hash does not match the stored identifier.
  \item Allocate a fresh VM ID for the new virtual machine instance.
  \item Execute \textsc{Identity}: run the capsule code, which establishes
        the word definitions and initial state specific to this capsule.
  \item Execute \textsc{Personality}: load block~1 from the kernel ramdrive
        (LBN~2048), which supplies the VM's behavioral personality.
  \item Log a parity record containing: the VM ID, the capsule hash, and
        the dictionary hash at the moment personality execution completes.
\end{enumerate}

Parity logging is implemented in \texttt{src/starkernel/vm/parity.c}.
Each birth and execution event is recorded with sufficient information
for an independent system to load the same capsule and compare dictionary
hashes. Zero deviation in dictionary hashes across independent runs
constitutes the primary evidence for deterministic behavior.

\section{Capsule Descriptor}
\label{vol2:sec:capsules:descriptor}

%% SOURCE: .claude/CLAUDE.md (CapsuleDesc)

Each capsule is described by a \texttt{CapsuleDesc} structure. The structure
is 64 bytes in size and cache-line aligned to avoid false sharing when the
capsule directory is scanned at birth time.

\begin{table}[ht]
  \centering
  \caption{CapsuleDesc fields}
  \label{tab:capsule-desc}
  \begin{tabular}{lll}
    \toprule
    Field & Type & Description \\
    \midrule
    content hash & \texttt{uint64\_t} & XXHash64 of capsule binary content \\
    type code & \texttt{uint8\_t} & \texttt{(m)}, \texttt{(p)}, or \texttt{(e)} \\
    name & \texttt{char[]} & Null-terminated capsule name \\
    \bottomrule
  \end{tabular}
\end{table}

%% TODO(bob): confirm the full field list for CapsuleDesc from include/vm.h
%% or starkernel/capsule/capsule_loader.c — the CLAUDE.md summary names
%% only content hash and type; the table above should be expanded with the
%% remaining fields (size, flags, block offset, etc.) once verified.

\section{Memory Layout}
\label{vol2:sec:capsules:memory}

%% SOURCE: .claude/CLAUDE.md (Kernel memory layout section)

The LithosAnanke kernel partitions VM memory as follows. Block RAM occupies
logical block numbers (LBN) 0 through 991, providing 1~MB of dedicated
RAM-backed blocks. The kernel ramdrive occupies LBN 2048 through 3071,
providing a further 1~MB used for capsule loading. LBN~2048 is the entry
point for \texttt{init.4th}: when the birth protocol executes Personality,
execution begins at this block. The log region occupies LBN 3072 through
5120, providing 2~MB for persistent logging at up to 32,768 lines of
64~bytes each. The full VM memory space is 5~MB in the hosted build
(\texttt{VM\_MEMORY\_SIZE}), with the first 2~MB reserved for the
dictionary (\texttt{DICTIONARY\_BLOCKS} = 2048 blocks).

\begin{table}[ht]
  \centering
  \caption{LithosAnanke LBN layout}
  \label{tab:lbn-layout}
  \begin{tabular}{lll}
    \toprule
    LBN range & Size & Purpose \\
    \midrule
    0 -- 991      & 1 MB & Block RAM (dedicated RAM blocks) \\
    2048 -- 3071  & 1 MB & Kernel ramdrive; LBN 2048 = \texttt{init.4th} entry \\
    3072 -- 5120  & 2 MB & Persistent log (32,768 lines $\times$ 64 bytes) \\
    \bottomrule
  \end{tabular}
\end{table}

\section{Capsule Files}
\label{vol2:sec:capsules:files}

%% SOURCE: .claude/CLAUDE.md (Capsule files in capsules/)
%% SOURCE: ls /home/user/StarForth/capsules/

The \texttt{capsules/} directory contains 19 \texttt{.4th} source files.
These are assembled by \texttt{tools/mkcapsule.c} into the binary capsule
directory format, which is compiled into the kernel image as
\texttt{capsule\_generated.c}.

\begin{table}[ht]
  \centering
  \caption{Capsule \texttt{.4th} source files}
  \label{tab:capsule-files}
  \begin{tabular}{ll}
    \toprule
    File & Purpose \\
    \midrule
    \texttt{init.4th}            & Default Mama VM personality \\
    \texttt{init-0.4th}          & Numbered personality variant 0 \\
    \texttt{init-1.4th}          & Numbered personality variant 1 \\
    \texttt{init-2.4th}          & Numbered personality variant 2 \\
    \texttt{init-3.4th}          & Numbered personality variant 3 \\
    \texttt{init-4.4th}          & Numbered personality variant 4 \\
    \texttt{init-5.4th}          & Numbered personality variant 5 \\
    \texttt{init-6.4th}          & Numbered personality variant 6 \\
    \texttt{init-7.4th}          & Numbered personality variant 7 \\
    \texttt{init-8.4th}          & Numbered personality variant 8 \\
    \texttt{init-9.4th}          & Numbered personality variant 9 \\
    \texttt{init-l8-stable.4th}  & L8 Jacquard: stable mode \\
    \texttt{init-l8-volatile.4th}& L8 Jacquard: volatile mode \\
    \texttt{init-l8-diverse.4th} & L8 Jacquard: diverse mode \\
    \texttt{init-l8-temporal.4th}& L8 Jacquard: temporal mode \\
    \texttt{init-l8-transition.4th} & L8 Jacquard: transition mode \\
    \texttt{init-l8-omni.4th}    & L8 Jacquard: omni mode \\
    \texttt{ACL.4th}             & Word-level ACL subsystem (self-activating) \\
    \texttt{zuse.4th}            & Bootstrap superuser; loaded by \texttt{ACL.4th} \\
    \bottomrule
  \end{tabular}
\end{table}

The \texttt{mkcapsule} tool (\texttt{tools/mkcapsule.c}) reads each
\texttt{.4th} source file, computes its XXHash64 content hash, and writes
a binary capsule directory entry. The resulting \texttt{capsule\_generated.c}
is a C translation unit that, when compiled into the kernel image, provides
the capsule directory accessible at boot time without filesystem access.

\section{Mama Forth Supervisor}
\label{vol2:sec:capsules:mama}

%% SOURCE: docs/formal/scraps/architecture/MAMA_FORTH_SUPERVISOR_ARCHITECTURE.tex

Mama Forth is the governing FORTH VM that orchestrates the multi-VM
ecosystem. It is itself a FORTH VM, a first-class citizen carrying the same
physics instrumentation as every VM it governs. The supervisor does not
stand outside the runtime model; it participates in it.

The ecosystem forms a tree. Mama Forth sits at the root. Beneath it, child
VMs each carry their own physics state. Beneath each child, block devices
carry per-device physics. Mama observes all three layers and makes
system-level decisions: when to spawn or kill a child VM, how to allocate
block devices to children, how to enforce governance across the ecosystem,
and how to respond to anomalous conditions.

The central design insight is that physics metadata becomes a three-tier
hierarchy:

\begin{itemize}
  \item \textbf{Word level} --- per-word metrics: execution heat, temperature,
        latency.
  \item \textbf{VM level} --- per-VM aggregates: child health, resource usage.
  \item \textbf{Block level} --- per-block device metrics: I/O patterns,
        thermal distribution.
\end{itemize}

Mama runs with elevated privileges along three axes: observing all word
executions in children, all block-device I/O, all child lifecycle events,
and all errors; controlling child VM spawn and kill, block allocation and
rebalancing; and coordinating shared block resources and load balancing.

%% TODO(bob): Confirm whether Mama Forth runs as a dedicated thread, a
%% coroutine, or the main loop. Source material leaves the concurrency
%% model open. The MAMA_FORTH_SUPERVISOR_ARCHITECTURE scrap notes this
%% is unresolved.

The Mama VM is initialized by the \texttt{MAMA\_INIT} capsule (type code
\texttt{(m)}). Exactly one such capsule exists per deployment. Child VMs
are born from \texttt{PRODUCTION} or \texttt{EXPERIMENT} capsules via
the birth protocol described in Section~\ref{vol2:sec:capsules:birth}.
